---
title: spatsoc paper 2
author:
  - name: Alec L. Robitaille
    orcid: 0000-0002-4706-1762
    corresponding: true
    email: robit.alec@gmail.com
    # roles:
    affiliations:
      - Memorial University of Newfoundland and Labrador
  - name: Quinn M. R. Webber
    orcid: 0000-0002-0434-9360
    corresponding: false
    # roles: []
    affiliations:
      - University of Guelph
  - name: Eric Vander Wal
    orcid: 0000-0002-8534-4317
    corresponding: false
    affiliations:
      - Memorial University of Newfoundland and Labrador
keywords:
  - R
  - social dynamics
  - collective movement
abstract: | 
  1. Beyond social networks, dyadic measures
  2. Extending social networks derived from GPS using detailed dyadic measures beyond rates of association
  3. Linking social processes to spatial/geography environment, hint at selection
  4. Expanding spatsoc to version N by including important and unavailable methods in R 
plain-language-summary: |
  New functions for the R package {spatsoc} to measure animal intragroup social dynamics
# key-points:
date: last-modified
bibliography: references/references.bib
# citation:
  # container-title: 
number-sections: true
---

<!-- TODO: abstract above -->

# Introduction

```{r}
#| label: setup
#| include: false
targets::tar_source('R')
```

Animal social dynamics have broad implications in ecology including resource
availability, disease transmission and collective behaviour. Social network
analysis can measure the structure of social systems, the consequences of an
individual's position in a network, the spread of disease or information across
networks, and the influence of the environment [@Farine_2015]. Open source tools
have been developed to derive social networks from animal telemetry data
[@Robitaille_2019] using aggregated association rates or home range overlap. The
methods have been used to measure contact rates to measure social cohesion
[@Bracken_2022], human-wildlife conflict [@Boudreau_2022], socioecology and
resource availability [@Peignier_2019],  community structure [@Sunga_2021],
influence of stress of social proximity [@Keshavarzi_2023], social patterns in
non-social species [@Heeres_2024].

Many facets of group living need to be measured to better understand animal
social systems [@King_2018]. There are additional types of dyadic relationships
including genetic, affiliative, agonistic, and cooperative [@Farine_2015]. There
are also interactions between spatial and social phenotypes and environments
[@Webber_2023]. Using direct observations, researchers have studied intragroup
social dynamics including social foraging [@Barnard_1981; @Giraldeau_1986;
@Hirsch_2020], spatial position within groups [@Krause_1994; @Monus_2008] and
the influence of individual traits such as age, sex and dominance
[@Furuichi_1983; @King_2009; @Teichroeb_2015]. Unfortunately, direct
observations can be labor-intensive to collect and susceptible to observer bias
[@Smith_2021], with low spatial and temporal resolution making them imperfect
for studies of collective movement [@King_2018].

There are expanded opportunities to understand animal social systems by pairing
advances in remote tracking technology with emerging approaches to measuring
intragroup social dynamics. The availability and resolution of remote tracking
data continues to improve [@Kays_2015; @Kays_2022; @Nathan_2022]. There are over
two decades of research on measuring the interaction between moving individuals
from the study of "moving point objects" in GISciences [@Andersson_2008;
@Dodge_2008; @Laube_2005] to group coordination of moving groups [@Couzin_2002;
@Herbert-Read_2011] and leadership and decision making [@King_2011; @Nagy_2010;
@Strandburg-Peshkin_2015]. Despite this, we suggest the availability of open
source tools [@Hampton_2015; @Powers_2019; @Wilson_2017] is the missing piece
between the call to integrate technology enhanced research with
multidisciplinary approaches to better understand animal social systems
[@Couzin_2023; @King_2018].

Here we present 
`r length(grep('cent|dir|dista|ali|del|zon|lead|get_geo|fus', ls('package:spatsoc')))`
new functions for the R package {spatsoc} [@Robitaille_2019] for measuring
intragroup dynamics. We conducted a scoping review to identify metrics
that were  applicable to spatiotemporal data, and focused on dyadic
interactions, dominance hierarchies and leadership patterns. Of the
resulting metrics, we selected those with a high number of citations (as a
measure of value for the target audience) and lack of availability in the R
programming language, either through R packages or supplemental materials in the
literature. We will demonstrate these new functions through a case study and
discuss biological interpretation.

<!-- TODO: update count if include randomization method --> 

# Methods

We conducted a scoping review to identify metrics used in ecology studies for
measuring dyadic interactions, dominance hierarchies and leadership patterns. We
focused on metrics applicable to spatiotemporal data such as animal movement
data. See details in Supplement A: Scoping Review Methods and Supplement B:
Scoping Review Results. Metrics were selected for inclusion in {spatsoc} based
on their relative importance as measured by frequency of use in the review, as
well as their lack of availability in R packages when the review was conducted.
Some metrics were specifically not chosen for inclusion such as metrics relating
to posture, metrics relating to static interaction or home range overlap and
metrics based on statistical models such as Hidden Markov model and Granger
causality. A full list of metrics not selected is in Supplement C: Selected
Metrics.

<!-- TODO: table of new functions, brief description, "family" -->

<!--
For each

- Citation
- Interpretation, **emphasis on biology**
- Formula
- Code example
- Similar terms? eg. see metric synonyms

See list of not selected metrics (Supplement  B), including already available 
functions in eg. {wildlifeDI}, metrics relating to posture,
static interaction eg. home range overlap, ...

-->


# Case study

Functionality previously available in {spatsoc} includes spatiotemporal
grouping, distance based edge-list generation and data-stream permutations
[@Robitaille_2019]. These functions have allowed users to detect interactions
between individuals, measure co-occurrence within and across species, and
generate social networks from telemetry data.
<!-- (TODO: cite examples citing spatsoc) -->

Spatiotemporal grouping can be performed given animal movement data, temporal
and spatial thresholds, and one of three grouping functions: `group_pts`,
`group_lines` or `group_polys`. These return a unique identifier for each
spatiotemporal group defined by point-based distances, linear trajectory
overlap and home range overlap, respectively.

`get_geometry` is a new helper function added to {spatsoc} to provide an
alternative interface for providing coordinates and, optionally, coordinate
reference system. Instead of providing the names to columns for the X and Y
coordinates using the `coords` argument, the `get_geometry` function adds a
simple feature geometry list column (`sfc`) using the {sf} package
[@Pebesma_2018]. The geometry method is used by default for all {spatsoc}
functions requiring coordinates when the `coords` argument is left null. This
interface also extends {spatsoc} to allow users to provide coordinates in any of
the many coordinate reference systems, both projected and longlat degrees, that
the {sf} package supports. The following examples and case study will illustrate
the geometry interface throughout.

For example, the suggested `group_pts` workflow with the example data from the
package and the `get_geometry()` helper function:

<!-- TODO: note get_geometry --> 

```{r}
#| label: calc-group-pts
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Set variables
temporal_threshold  <- '20 minutes'
spatial_threshold <- 50
id <- 'ID'
coords <- c('X', 'Y')
utm <- 32736

# Setup DT for geometry interface
get_geometry(DT, coords = coords, crs = utm)

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = temporal_threshold)

# Spatial grouping with timegroup
group_pts(DT, threshold = spatial_threshold, id = id,
          coords = coords, timegroup = 'timegroup')
```

```{r}
#| label: tbl-group-pts
#| tbl-cap: "Spatiotemporal groups measured using group_times and group_pts. group_times returns a 'timegroup' column indicating temporal groups within the specified temporal threshold. group_pts returns a 'group' column indicating spatiotemporal groups within the provided timegroup and specific distance threshold."
#| echo: false
DT <- tar_read(spatial_groups)
sel_groups <- DT[ID != 'A'][, .N, by = group][between(N, 2, 3), sample(group, 3)]
tinytable::tt(DT[group %in% sel_groups][order(timegroup)][
  seq.int(6), .(ID, datetime, geometry, timegroup, group)])
```

Users can also measure interindividual distance using the `edge_dist` function
and identify nearest neighbors using the `edge_nn` function. For example:

```{r}
#| label: calc-edge-dist
#| echo: true
#| eval: false
interindividual_dist <- edge_dist(
  DT,
  threshold = spatial_threshold,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)
dyad_id(interindividual_dist, id1 = 'ID1', id2 = 'ID2')
```

```{r}
#| label: tbl-edge-dist
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_dist. Using the 'timegroup' column returned by group_times, edge_dist returns a distance based edge-list where individuals in the column 'ID1' are within the provided distance threshold from 'ID2'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and NAs can be returned where an individual is not within the provided distance threshold if 'fillNA' is TRUE."
tar_load(distance_edges)
sel_tg <- distance_edges[, .N, by = .(timegroup, is.na(ID2))][, .N, timegroup][N > 1][, sample(timegroup, 1)]
tinytable::tt(distance_edges[timegroup == sel_tg][seq.int(6)])
```

```{r}
#| label: calc-edge-nn
#| echo: true
#| eval: false
nearest_neighbours <- edge_nn(
  DT,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE
)
dyad_id(nearest_neighbours, id1 = 'ID1', id2 = 'ID2')
```

```{r}
#| label: tbl-edge-nn
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_nn. Using the 'timegroup' column returned by group_times, edge_nn returns a distance based edge-list where individuals in the column 'ID2' are the nearest neighbour to focal individuals in the column 'ID1'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and a distance threshold can be provided using 'threshold'."
tar_load(nn_edges)
sel_tg <- nn_edges[, .N, by = timegroup][N > 1][, sample(timegroup, 1)]
tinytable::tt(nn_edges[timegroup == sel_tg][seq.int(6)])
```

Building on these, {spatsoc}'s new functions provides a finer understanding of
an individual's behaviour with respect to conspecifics. We aim to provide users
with accessible, flexible functions that will help to identify leader follower
patterns, fission fusion dynamics and potentially dominance behaviours from
animal telemetry data in their ecological systems.

## Position within groups

Extending {spatsoc}'s abilities to identify spatiotemporal groups, a new set of
functions are now available to measure individuals position relative within
spatiotemporal groups. After identifying spatiotemporal groups with eg.
`group_pts`, we can measure the group centroid, as defined by the mean of
individual locations in a group.

```{r}
#| label: calc-cent-group
#| eval: false
#| echo: true
centroid_group(DT)
```

```{r}
#| label: tbl-cent-group
#| eval: true
#| echo: false
#| tbl-cap: "Group centroids measured using centroid_group. Using the 'group' column returned by group_pts, centroid_group returns the column 'centroid' representing the centroid of all individual's locations in each spatiotemporal group."
tar_load(group_centroids)
tinytable::tt(
  group_centroids[
    1:5, 
    .(ID, datetime, timegroup, group, geometry, centroid)]
)
```

Then we can measure each individual's direction and distance to the group
centroid. The distance to group centroid is the geographic distance from the
focal individual to the group centroid. The direction to group centroid is the
absolute direction from the focal individual to the group centroid. The rank
distance to group centroid is the ordinal rank of distances to the group
centroid within each spatiotemporal group. Spatial position within groups 
with respect to the group centroid has been studied in relation to 
predation risk [@Heesen_2015], sex differences [@Inoue_2019], and leadership
[@Jolles_2017; @Kano_2021].

```{r}
#| label: calc-dist-dir-cent
#| echo: true
#| eval: false
distance_to_centroid(DT, return_rank = TRUE)
direction_to_centroid(DT)
```

```{r}
#| label: fig-dist-dir-cent
#| echo: false
#| eval: true
#| warning: false
#| fig-cap: "A) Distance and B) direction to group centroids measured using distance_to_centroid and direction_to_centroid. Using the 'centroid' column returned by centroid_group, distance_to_centroid and direction_to_centroid return the columns 'distance_centroid' and 'direction_centroid' representing the focal individual's distance and direction to the group centroid. Optionally, the rank distance to the group centroid can be returned by distance_to_centroid if 'return_rank' is true."
tar_read(fig_dist_dir_cent)
```

## Position relative to leader

After considering how individuals are positioned relative to the group centroid,
we can also measure how individuals are positioned relative to the mean group
direction. To do so, we rotate the coordinate system around the group centroid
by the mean direction of all individuals in the group and take the distance
along this new axis as the measure of front-back position within the group
[@Quera_2023; @Harel_2021; @Andrienko_2013]. The rank position within group is
the ordinal rank along the front-to-back axis of the group's direction
[@Burns_2012]. 

```{r}
#| label: calc-leader-dir-group
#| eval: false
#| echo: true
direction_step(DT, id = id)
leader_direction_group(DT, return_rank = TRUE)
```

```{r}
#| label: fig-leader-dir-group
#| eval: true
#| echo: false
#| fig-cap: "A) Example spatiotemporal group showing relative positions of individuals with respect to the group centroid and mean group direction. B) Position along the group direction measured using leader_direction_group. Using the 'direction' column returned by direction_step and 'group' column returned by group_pts, leader_direction_group returns a column named 'position_group_direction' indicating the front-back position along the mean group direction. Optionally, the rank position along the mean group direction if 'return_rank' is true."
tar_read(fig_pos_group_dir)
```

Taking this simple, dynamic definition of leadership, we can identify the
distance and direction of individuals to the leader of each spatiotemporal
group. The distance to leader is the geographic distance from the focal
individual to the group's leader and the direction to leader is the absolute
direction from the focal individual to the group's leader.

```{r}
#| label: calc-dist-dir-lead
#| eval: false
#| echo: true
distance_to_leader(DT)
direction_to_leader(DT)
```

```{r}
#| label: fig-dist-dir-lead
#| eval: true
#| echo: false
#| fig-cap: "A) Distance and B) direction to group leaders measured using distance_to_leader and direction_to_leader. Using the 'position_group_direction' column returned by leader_direction_group, distance_to_leader and direction_to_leader return columns named 'distance_leader' and 'direction_leader' indicating the distance and direction of the focal individual to the leader of the spatiotemporal group."
tar_read(fig_dist_dir_leader)
```



## Fission fusion

Given the variability in defining fission fusion dynamics in the literature
(Supplement B: Scoping Review Results), we developed a flexible function that
allows users to both use their system specific definitions and at the same time
easily compare to results to other definitions.

<!-- TODO: add table numbers and cite them directly eg. Tab SB-3 instead of 
Supplement B: .... Table -->

Function arguments: 

- `threshold`: spatial distance threshold to establish a fusion event
- `n_min_length`: the minimum number of successive fixes that are required to 
establish a fusion event
- `n_max_missing`: the maximum number of allowable missed observations for 
either individual in a dyad within a fusion event
- `allow_split`: if fusion events allow a temporary spatial splitting for one
observation without resulting in a fission event

<!-- See Table Sup B: varying definitions of fission fusion 
TODO: mention from edge_dist
-->

```{r}
#| label: calc-fusion
#| eval: false
#| echo: true
fusion_events <- fusion_id(
  interindividual_dist,
  threshold = spatial_threshold, 
  n_min_length = 3, 
  n_max_missing = 1, 
  allow_split = TRUE
)
```


```{r}
#| label: fig-fusion
#| echo: false
#| eval: true
#| fig-cap: "A) Example locations for individuals 'A' and 'C' showing fusion event lasting three observations. B) A fusion event identified using fusion_id. Using the distance based edge-lists measured using edge_dist and the dyad id returned by dyad_id, fusion_id identifies fusion events according to user specified constraints: the distance threshold ('threshold'), the minimum number of observations defining a fusion event ('n_min_length'), the maximum number of missing observations from either individual in a fusion event ('n_max_missing') and if a single observation where individuals are beyond the distance threshold should result in a fission event ('allow_split'). In B), a fusion event is identified for individuals A and C given they are within the distance threshold at timegroups 2, 3, and 5, and 'allow_split' is true."
tar_read(fig_fusion_events)
```


## Directions

Next, we can use individuals' movement directions to explore group polarization, interindividual direction, and
directional alignment.

<!-- TODO: describe spatial measures -->

<!-- TODO: references -->
<!-- related eg. direction between individuals (always L / R), polarization of flock, -->
<!-- etc -->

Polarization measures the uniformity of absolute direction in a group of
individuals [@Wang_2022; @Jolles_2017] on a scale of 0-1 where values near 0
indicate that individuals are moving in completely different directions (or
"non-aligned") and values near 1 indicate that individuals are moving in similar
directions or ("aligned").

```{r}
#| label: calc-dir-pol
#| echo: true
#| eval: false
direction_polarization(DT)
```

```{r}
#| label: fig-dir-pol
#| echo: false
#| eval: true
#| fig-cap: "Direction polarization measured using direction_polarization. Using the 'direction' column returned by direction_step and 'group' column returned by group_pts, direction_polarization returns a column named 'polarization' indicating the polarization in directions in each spatiotemporal group. Ranges of directions shown for a sample of groups in four equal interval categories of direction polarization. Values of polarization range from 0 (non-alignment) to 1 (alignment)."
tar_read(fig_dir_polarization)
```

Directional alignment is the relative difference between two individuals' 
directions. Given the similarity to `edge_dist`, this functionality
is provided under a similar name: `edge_alignment`. 

<!-- TODO: this similarity is edge-list generating -->

```{r}
#| label: calc-edge-align
#| echo: true
#| eval: false
directional_align <- edge_alignment(
  DT = DT,
  id = id,
  signed = FALSE
)
```

Interindividual direction measures the absolute direction between individuals. 
<!-- TODO: describe spatial measures -->

```{r}
#| label: calc-edge-dir
#| echo: true
#| eval: false
dyad_directions <- edge_direction(
  edges = interindividual_dist,
  DT = DT,
  id = id
)
```


```{r}
#| label: fig-edge-align-dir
#| echo: false
#| eval: true
#| fig-cap: ""
tar_read(fig_edge_dir_and_align)
```


## Lagged differences in direction

The directional correlation delay [@Nagy_2010] of individuals i, j is given by

$$C_{ij} = [\overrightarrow{v_{i}}(t) * \overrightarrow{j}(t + \tau)]_{t}$$

where 

- $\overrightarrow{v_{i}}(t)$ is the normalized velocity of individual i at time t
- $\overrightarrow{v_{j}}(t + \tau)$ is the normalized velocity of individual j at time t + $\tau$
- Note that $C_{ij}(\tau)$ = $C_{ji}(-\tau)$
- Calculated only where pairs of individuals were less than 100 m apart

The maximum value of the directional correlation function $C_{ij}$ is at
$C_{ij}(\tau^{*}_{ij})$ where $\tau^{*}_{ij}$ is the directional correlation
delay time. $\tau^{*}_{ij}$ values focus on the relationship in pairs of
individuals, ignoring hierarchy changes caused by other individuals.

- Note that $\tau^{*}_{ij}$ = $-\tau^{*}_{ji}$
- Negative values indicate that flight directional changes of individual
i falls behind that of individual j and therefore j is leading

Hierarchical networks can be generated using $\tau^{*}_{ij}$. 

```{r}
#| label: calc-edge-delay
#| eval: false
#| echo: true
delay <- edge_delay(
  edges = fusion_events,
  DT = DT,
  window = 3,
  id = id
)
```

```{r}
#| label: fig-edge-delay
#| eval: true
#| echo: false
#| fig-cap: ''
tar_read(fig_edge_delay)
```


<!-- TODO: example of hierarchical networks -->

## Behavioural zones

The behavioural zones metric assigns neighbours to three non-overlapping
behavioural zones [Couzin_2002]. The "zone of repulsion" is the minimum distance
around an individual within which neighbours are expected to move away to avoid
collisions. The "zone of orientation" is the next zone around an individual
beyond the "zone of repulsion" within which neighbours are expected to orient
themselves to the movement of their neighbours. The "zone of attraction" is the
farthest zone around an individual within which neighbours are expected to be
attracted to the position of the focal individual. Notably, there is a possibly
"blind volume" behind the individual representing the limits of their
perception.


<!-- figure adapted from Couzin et al 2002 --> 

```{r}
#| label: calc-edge-zones
#| eval: false
#| echo: true
zone_thresholds <- c(10, 25, 50)
zone_labels <- c('repulsion', 'orientation', 'attraction')
blind_volume <- 3
delay <- edge_zones(
  interindividual_dist,
  zone_thresholds = zone_thresholds,
  zone_labels = zone_labels,
  blind_volume = blind_volume
)
```

```{r}
#| label: fig-edge-zones
#| eval: true
#| echo: false
#| fig-cap: ''
tar_read(fig_edge_zones)
```

# Discussion

- social networks, history of spatsoc
- high number of studies in review with IID, NND



# Conclusion

<!-- Test model only conclusions with increase species, increase systems given flexible powerful freely available functions -->

<!-- Association is only one layer of the animal social system -->

# Box 1 {.unnumbered}

Producer-scrounger dynamics describe two strategies of social foragers:
“producers” are those who discover patches and “scroungers” are those who join
patches discovered by others. This concept can be applied to animal telemetry
data by evaluating the position of individuals within groups, the individual’s
arrival timing to forage patches and characteristics of forage patches.
Individuals balance predation risk and foraging success by adjusting their
spatial position within groups  (Teichroeb 2015, Hirsch 2019), and the foraging
strategy selected by an individual can be influenced by the size of forage
patches (Hansen 2016).

Leader-follower patterns in groups of moving animals can be studied both to
determine factors that causally influence leadership, and for the influence of
leader-follower strategies on foraging success and downstream fitness. Applied
to animal telemetry, leadership can be defined simply by identifying the
individual at the foremost position along the front-to-back axis of the group’s
direction (Quera 2023, Harel 2021). Alternatively, leadership can be identified by
calculating the  directional correlation delay (Nagy 2010), the temporal delay
of shared directional movements between individuals. Leadership in a group of
moving animals can inform animal decision making, group coordination, and
migratory patterns (Flack 2018, Weesner 2023, Quera 2023, Kano 2021).

Individual traits can influence behaviors related to intragroup dynamics.
Dominance is related to foraging success (Caraco 1989, Teichroeb 2015) and
choice of foraging strategy (King 2009, Aplin and Morand-Ferron 2017, Lee 2016).
Dominant, aggressive, grooming and other interindividual behaviors are typically
measured using focal observations.  However, some metrics have been developed
for remotely estimating dominance behaviors. The directional correlation delay
can be used to derive hierarchies of leadership which can be a proxy for
dominance relationships in certain species (Flack 2013, Akos 2014). Movement
patterns that estimate dominance relationships can be identified from animal
telemetry including movement initiations (Amornbunchornvej 2018),
approach-avoidance behaviours (Strandburg-Peshkin 2015) and displacement events
(Evans 2018).



# References {.unnumbered}

:::{#refs}

:::


{{< include supplement/supplement-a.qmd >}}

{{< include supplement/supplement-b.qmd >}}

{{< include supplement/supplement-c.qmd >}}