---
title: spatsoc paper 2
author:
  - name: Alec L. Robitaille
    orcid: 0000-0002-4706-1762
    corresponding: true
    email: robit.alec@gmail.com
    # roles:
    affiliations:
      - Memorial University of Newfoundland and Labrador
  - name: Quinn M. R. Webber
    orcid: 0000-0002-0434-9360
    corresponding: false
    # roles: []
    affiliations:
      - University of Guelph
  - name: Eric Vander Wal
    orcid: 0000-0002-8534-4317
    corresponding: false
    affiliations:
      - Memorial University of Newfoundland and Labrador
keywords:
  - R
  - social dynamics
  - collective movement
abstract: | 
  1. Beyond social networks, dyadic measures
  2. Extending social networks derived from GPS using detailed dyadic measures beyond rates of association
  3. Linking social processes to spatial/geography environment, hint at selection
  4. Expanding spatsoc to version N by including important and unavailable methods in R 
plain-language-summary: |
  New functions for the R package {spatsoc} to measure animal intragroup social dynamics
# key-points:
date: last-modified
bibliography: references/references.bib
# citation:
  # container-title: 
number-sections: true
---

<!-- TODO: abstract above -->

# Introduction

```{r}
#| label: setup
#| include: false
targets::tar_source('_targets.R')
```

Animal social dynamics have broad implications in ecology including resource
availability, disease transmission and collective behaviour. Social network
analysis can measure the structure of social systems, the consequences of an
individual's position in a network, the spread of disease or information across
networks, and the influence of the environment [@Farine_2015]. Open source tools
have been developed to derive social networks from animal telemetry data
[@Robitaille_2019] using aggregated association rates or home range overlap. The
methods have been used to measure contact rates to measure social cohesion
[@Bracken_2022], human-wildlife conflict [@Boudreau_2022], socioecology and
resource availability [@Peignier_2019],  community structure [@Sunga_2021],
influence of stress of social proximity [@Keshavarzi_2023], social patterns in
non-social species [@Heeres_2024].

Many facets of group living need to be measured to better understand animal
social systems [@King_2018]. There are additional types of dyadic relationships
including genetic, affiliative, agonistic, and cooperative [@Farine_2015]. There
are also interactions between spatial and social phenotypes and environments
[@Webber_2023]. Using direct observations, researchers have studied intragroup
social dynamics including social foraging [@Barnard_1981; @Giraldeau_1986;
@Hirsch_2020], spatial position within groups [@Krause_1994; @Monus_2008] and
the influence of individual traits such as age, sex and dominance
[@Furuichi_1983; @King_2009; @Teichroeb_2015]. Unfortunately, direct
observations can be labor-intensive to collect and susceptible to observer bias
[@Smith_2021], with low spatial and temporal resolution making them imperfect
for studies of collective movement [@King_2018].

There are expanded opportunities to understand animal social systems by pairing
advances in remote tracking technology with emerging approaches to measuring
intragroup social dynamics. The availability and resolution of remote tracking
data continues to improve [@Kays_2015; @Kays_2022; @Nathan_2022]. There are over
two decades of research on measuring the interaction between moving individuals
from the study of "moving point objects" in GISciences [@Andersson_2008;
@Dodge_2008; @Laube_2005] to group coordination of moving groups [@Couzin_2002;
@Herbert-Read_2011] and leadership and decision making [@King_2011; @Nagy_2010;
@Strandburg-Peshkin_2015]. Despite this, we suggest the availability of open
source tools [@Hampton_2015; @Powers_2019; @Wilson_2017] is the missing piece
between the call to integrate technology enhanced research with
multidisciplinary approaches to better understand animal social systems
[@Couzin_2023; @King_2018].

Here we present 
`r length(grep('cent|dir|dista|ali|del|zon|lead|get_geo|fus', ls('package:spatsoc')))`
new functions for the R package {spatsoc} [@Robitaille_2019] for measuring
intragroup dynamics. We conducted a scoping review to identify metrics
that were  applicable to spatiotemporal data, and focused on dyadic
interactions, dominance hierarchies and leadership patterns. Of the
resulting metrics, we selected those with a high number of citations (as a
measure of value for the target audience) and lack of availability in the R
programming language, either through R packages or supplemental materials in the
literature. We will demonstrate these new functions through a case study and
discuss biological interpretation.

<!-- TODO: update count if include randomization method --> 

# Methods

We conducted a scoping review to identify metrics used in ecology studies for
measuring dyadic interactions, dominance hierarchies and leadership patterns. We
focused on metrics applicable to spatiotemporal data such as animal movement
data. See details in Supplement A: Scoping Review Methods and Supplement B:
Scoping Review Results. Metrics were selected for inclusion in {spatsoc} based
on their relative importance as measured by frequency of use in the review, as
well as their lack of availability in R packages when the review was conducted.
Some metrics were specifically not chosen for inclusion such as metrics relating
to posture, metrics relating to static interaction or home range overlap and
metrics based on statistical models such as Hidden Markov model and Granger
causality. A full list of metrics not selected is in Supplement C: Selected
Metrics.

<!-- TODO: table of new functions, brief description, "family" -->

<!--
For each

- Citation
- Interpretation, **emphasis on biology**
- Formula
- Code example
- Similar terms? eg. see metric synonyms

See list of not selected metrics (Supplement  B), including already available 
functions in eg. {wildlifeDI}, metrics relating to posture,
static interaction eg. home range overlap, ...

-->


# Case study

Functionality previously available in {spatsoc} includes spatiotemporal
grouping, distance based edge-list generation and data-stream permutations
[@Robitaille_2019]. These functions have allowed users to detect interactions
between individuals, measure co-occurrence within and across species, and
generate social networks from telemetry data.
<!-- (TODO: cite examples citing spatsoc) -->

Spatiotemporal grouping can be performed given animal movement data, temporal
and spatial thresholds, and one of three grouping functions: `group_pts()`,
`group_lines()` or `group_polys()`. These return a unique identifier for each
spatiotemporal group defined by point-based distances, linear trajectory
overlap and home range overlap, respectively.

`get_geometry()` is a new helper function added to {spatsoc} to provide an
alternative interface for providing coordinates and, optionally, coordinate
reference system. Instead of providing the names to columns for the X and Y
coordinates using the `coords` argument, the `get_geometry()` function adds a
simple feature geometry list column (`sfc`) using the {sf} package
[@Pebesma_2018]. The geometry method is used by default for all {spatsoc}
functions that require coordinates when the `coords` argument is left null. This
interface also extends {spatsoc} to allow users to provide coordinates in any of
the many coordinate reference systems, both projected and longlat degrees, that
the {sf} package supports. The following examples and case study will illustrate
the geometry interface throughout.

For example, the suggested `group_pts()` workflow with the example data from the
package and the `get_geometry()` helper function:

<!-- TODO: note get_geometry --> 

```{r}
#| label: calc-group-pts
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Set variables
temporal_threshold  <- '20 minutes'
spatial_threshold <- 50
id <- 'ID'
coords <- c('X', 'Y')
utm <- 32736

# Setup DT for geometry interface
get_geometry(DT, coords = coords, crs = utm)

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = temporal_threshold)

# Spatial grouping with timegroup
group_pts(DT, threshold = spatial_threshold, id = id,
          coords = coords, timegroup = 'timegroup')
```

```{r}
#| label: tbl-group-pts
#| tbl-cap: "Spatiotemporal groups measured using group_times() and group_pts(). group_times() returns a 'timegroup' column indicating temporal groups within the specified temporal threshold. group_pts() returns a 'group' column indicating spatiotemporal groups within the provided timegroup and specific distance threshold. Example output showing a subset of three spatiotemporal groups."
#| echo: false
DT <- tar_read(spatial_groups)
sel_groups <- DT[ID != 'A'][, .N, by = group][between(N, 2, 3), sample(group, 3)]
tinytable::tt(DT[group %in% sel_groups][order(timegroup)][
  seq.int(6), .(ID, datetime, geometry, timegroup, group)])
```

Users can also measure interindividual distance using the `edge_dist()` function
and identify nearest neighbors using the `edge_nn()` function. For example:

```{r}
#| label: calc-edge-dist
#| echo: true
#| eval: false
interindividual_dist <- edge_dist(
  DT,
  threshold = spatial_threshold,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)
dyad_id(interindividual_dist, id1 = 'ID1', id2 = 'ID2')
```

```{r}
#| label: tbl-edge-dist
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_dist(). Using the 'timegroup' column returned by group_times(), edge_dist() returns a distance based edge-list where individuals in the column 'ID1' are within the provided distance threshold from 'ID2'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and NAs can be returned where an individual is not within the provided distance threshold if 'fillNA' is TRUE. Example output showing six observations of one timegroup with distances returned between focal individual 'ID1' and neighbour 'ID2'."
tar_load(distance_edges)
sel_tg <- distance_edges[, .N, by = .(timegroup, is.na(ID2))][, .N, timegroup][N > 1][, sample(timegroup, 1)]
tinytable::tt(distance_edges[timegroup == sel_tg][seq.int(6)])
```

```{r}
#| label: calc-edge-nn
#| echo: true
#| eval: false
nearest_neighbours <- edge_nn(
  DT,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE
)
dyad_id(nearest_neighbours, id1 = 'ID1', id2 = 'ID2')
```

```{r}
#| label: tbl-edge-nn
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_nn(). Using the 'timegroup' column returned by group_times(), edge_nn() returns a distance based edge-list where individuals in the column 'ID2' are the nearest neighbour to focal individuals in the column 'ID1'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and a distance threshold can be provided using 'threshold'. Example output showing six observations of one timegroup with distances returned between focal individual 'ID1' and nearest neighbour 'ID2'."
tar_load(nn_edges)
sel_tg <- nn_edges[, .N, by = timegroup][N > 1][, sample(timegroup, 1)]
tinytable::tt(nn_edges[timegroup == sel_tg][seq.int(6)])
```

Building on these, {spatsoc}'s new functions provides a finer understanding of
an individual's behaviour with respect to conspecifics. We aim to provide users
with accessible, flexible functions that will help to identify leader follower
patterns, fission fusion dynamics and potentially dominance behaviours from
animal telemetry data in their ecological systems.

## Position within groups

Extending {spatsoc}'s abilities to identify spatiotemporal groups, a new set of
functions are now available to measure individuals position relative within
spatiotemporal groups. After identifying spatiotemporal groups with eg.
`group_pts()`, we can measure the group centroid, as defined by the mean of
individual locations in a group, using `centroid_group()`.

```{r}
#| label: calc-cent-group
#| eval: false
#| echo: true
centroid_group(DT)
```

```{r}
#| label: tbl-cent-group
#| eval: true
#| echo: false
#| tbl-cap: "Group centroids measured using centroid_group(). Using the 'group' column returned by group_pts(), centroid_group() returns the column 'centroid' representing the centroid of all individual's locations in each spatiotemporal group. Example output showing a subset of the centroids for three spatiotemporal groups."
tar_load(group_centroids)
sel_groups <- group_centroids[ID != 'A'][, .N, by = group][between(N, 2, 3), sample(group, 3)]
tinytable::tt(group_centroids[group %in% sel_groups][order(timegroup)][
  seq.int(6), .(ID, timegroup, group, geometry, centroid)])
```

Then we can measure each individual's distance and direction to the group
centroid using `distance_to_centroid()` and `direction_to_centroid()`. The
distance to group centroid is the geographic distance from the focal individual
to the group centroid. The direction to group centroid is the direction
from the focal individual to the group centroid. The rank of individual's
distances to the group centroid can be optionally returned using the argument
`return_rank`. Spatial position within groups with respect to the group centroid
has been studied in relation to predation risk [@Heesen_2015;
@Herbert-Read_2017], leadership [@Jolles_2017; @Kano_2021] and group cohesion
[@Harel_2021; @Bracken_2022].

```{r}
#| label: calc-dist-dir-cent
#| echo: true
#| eval: false
distance_to_centroid(DT, return_rank = TRUE)
direction_to_centroid(DT)
```

```{r}
#| label: fig-dist-dir-cent
#| echo: false
#| eval: true
#| warning: false
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "A) Distance and B) direction to group centroids measured using distance_to_centroid() and direction_to_centroid(). Using the 'centroid' column returned by centroid_group, distance_to_centroid() and direction_to_centroid() return the columns 'distance_centroid' and 'direction_centroid' representing the focal individual's distance and direction to the group centroid. Optionally, the rank distance to the group centroid can be returned by distance_to_centroid() if 'return_rank' is true."
tar_read(fig_dist_dir_cent)
```

After considering how individuals are positioned relative to the group centroid,
we can also measure how individuals are positioned relative to the mean group
direction. To do so, we rotate the coordinate system around the group centroid
by the mean direction of all individuals in the group and take the distance
along this new axis as the measure of front-back position within the group
[@Quera_2023; @Harel_2021; @Andrienko_2013] using the function
`leader_direction_group()`. The rank of individual's positions within
spatiotemporal groups [@Burns_2012] can be optionally returned with the argument
`return_rank = TRUE`.

```{r}
#| label: calc-leader-dir-group
#| eval: false
#| echo: true
direction_step(DT, id = id)
leader_direction_group(DT, return_rank = TRUE)
```

```{r}
#| label: fig-leader-dir-group
#| eval: true
#| echo: false
#| fig-cap: "A) Example spatiotemporal group showing relative positions and movement directions of individuals (small points and grey arrows) with respect to the group centroid and mean group direction (large point and black arrow). B) Position along the group direction measured using leader_direction_group(). Using the 'direction' column returned by direction_step() and 'group' column returned by group_pts(), leader_direction_group() returns a column named 'position_group_direction' representing the front-back position along the mean group direction. Optionally, the rank position along the mean group direction if 'return_rank' is true."
tar_read(fig_pos_group_dir)
```

Taking this simple, dynamic definition of leadership, we can identify the
distance and direction of individuals to the leader of each spatiotemporal group
using `distance_to_leader()` and `direction_to_leader()`. The distance to leader
is the geographic distance from the focal individual to the group's leader and
the direction to leader is the direction from the focal individual to the
group's leader. The front-back position within the group, and distance and
direction to the group leader has been used to study interaction rules
[@Weesner_2023], leadership [@Quera_2023; @Sarova_2010], risk avoidance
[@Josephs_2016], and how reproduction influences social structure
[@Perez-Barberia_2018].

```{r}
#| label: calc-dist-dir-lead
#| eval: false
#| echo: true
distance_to_leader(DT)
direction_to_leader(DT)
```

```{r}
#| label: fig-dist-dir-lead
#| eval: true
#| echo: false
#| fig-cap: "A) Distance and B) direction to group leaders measured using distance_to_leader and direction_to_leader. Using the 'position_group_direction' column returned by leader_direction_group(), distance_to_leader() and direction_to_leader() return columns named 'distance_leader' and 'direction_leader' indicating the distance and direction of the focal individual to the leader of the spatiotemporal group. Note: for individuals identified as leader using 'position_group_direction', the distance is 0 and the direction is NA."
tar_read(fig_dist_dir_leader)
```

## Directions

Next, we can use an individual's movement direction and the direction to
neighbors to explore polarization, interindividual direction, and directional
alignment. We can measure polarization, the uniformity of movement directions in
a group of individuals, using `direction_polarization()`. Polarization is
represented on a scale of 0-1 where values near 0 indicate that individuals in a
group are moving in completely different directions (or "non-aligned") and
values near 1 indicate that individuals are moving in similar directions or
("aligned"). Polarization in the movement direction of groups of animals has
been used to study cognition [@Wang_2022], learning [@Vega-Trejo_2020], the
influence of habitat [@Strandburg-Peshkin_2017] and individual differences
[@Jolles_2017] in collective behaviour.

```{r}
#| label: calc-dir-pol
#| echo: true
#| eval: false
direction_polarization(DT)
```

```{r}
#| label: fig-dir-pol
#| echo: false
#| eval: true
#| fig-cap: "Direction polarization measured using direction_polarization(). Using the 'direction' column returned by direction_step() and 'group' column returned by group_pts(), direction_polarization() returns a column named 'polarization' indicating the polarization in directions in each spatiotemporal group. Ranges of directions shown for a sample of groups in four equal interval categories of direction polarization. Values of polarization range from 0 (non-alignment) to 1 (alignment)."
tar_read(fig_dir_polarization)
```

The interindividual direction, or direction from a focal individual to a
neighbour, can be measured using `edge_direction()`. Like `edge_alignment()`,
the interindividual directions are returned by in an edge-list format.
Interindividual direction has been used to study risk exposure in schooling fish
[@Lemasson_2014], coordinated movement [@Herbert-Read_2016] and mass migration
[@Torney_2018].

```{r}
#| label: calc-edge-dir
#| echo: true
#| eval: false
dyad_directions <- edge_direction(
  edges = interindividual_dist,
  DT = DT,
  id = id
)
```

The directional alignment, or relative difference between two individuals'
directions, can be measured using `edge_alignment()`. The differences in
movement directions are returned in an edge-list format similar to `edge_dist()`
and other edge-list generating functions. Directional alignment has been used to
study leadership [@Sarova_2010], decision making [@Strandburg-Peshkin_2015], and
information transfer [@Sumpter_2018].

```{r}
#| label: calc-edge-align
#| echo: true
#| eval: false
directional_align <- edge_alignment(
  DT = DT,
  id = id,
  signed = FALSE
)
```


```{r}
#| label: fig-edge-align-dir
#| echo: false
#| eval: true
#| fig-cap: "A) Example locations for individuals 'E' (purple) and 'F' (blue) showing concurrent movement paths in timegroups 1, 2 and 3. B) Interindividual direction and directional_alignment() using edge_direction() and edge_alignment(). Using distance based edge-lists returned by edge_dist() or edge_nn() and the 'timegroup' column returned by `group_times()`, edge_direction() returns a column named 'direction_dyad' representing the direction between the focal individual ('ID1') and the neighboring individual ('ID2'). Using the 'direction' column returned by direction_step(), the edge_alignment() function returns a column named 'direction_diff' representing the absolute different in movement directions between two individuals ('ID1' and 'ID2'). Optionally, the signed difference can be returned by edge_alignment() if the argument 'signed' is true. "
tar_read(fig_edge_dir_and_align)
```

## Behavioural zones

The behavioural zones are non-overlapping regions around a focal individual
[@Couzin_2002] measured using `edge_zones()`. The distance from the focal
individual to each neighbour is used to assign the neighbours to behavioural
zones. In addition, the direction from the focal individual to each neighbour
can be used to determine if the neighbour is within a "blind volume" outside of
the individual's visual perception. @Couzin_2002 defined three non-overlapping
zones: 1) the "zone of repulsion" at a minimum distance around the focal
individual within which neighbours are expected to move away to avoid
collisions, 2) the "zone of orientation" around the focal individual beyond the
"zone of repulsion" within which neighbours are expected to orient themselves to
the movement of their neighbours, and 3) the "zone of attraction" beyond the
"zone of orientation" where neighbours are expected to e attracted to the
position of the focal individual. The `edge_zones()` is designed to allow for
flexible definitions of zones including distance thresholds, zone labels and
blind volume thresholds. The behavioural zones metric has been used to study the
impact of individual movement speed and predator evasion [@Klamser_2021] and
identifying rules of interaction in large animal groups [@Lukeman_2010].

```{r}
#| label: calc-edge-zones
#| eval: false
#| echo: true
zone_thresholds <- c(10, 25, 50)
zone_labels <- c('repulsion', 'orientation', 'attraction')
blind_volume <- 3

zones <- edge_zones(
  interindividual_dist,
  zone_thresholds = zone_thresholds,
  zone_labels = zone_labels,
  blind_volume = blind_volume
)
```

```{r}
#| label: fig-edge-zones
#| eval: true
#| echo: false
#| fig-cap: "Behavioural zones measured using edge_zones(). Using the 'distance' column returned by edge_dist(), the edge_zones() function returns the column 'zone' representing the behavioural zone of the neighbor (ID2) with respect to the focal individual (ID1). Optionally, the 'direction_dyad' column returned by edge_direction() can be used to define the blind volume, or range of interindividual directions that are beyond the visual perception of the focal individual."
tar_read(fig_edge_zones)
```


## Fission fusion

Given the variability in defining fission fusion dynamics in the literature
(Supplement B: Scoping Review Results: [Table @tbl-defn-fission]), the
`fusion_id()` function flexibly allows users to both use specific threshold and
definitions that match their study species and research questions and easily
compare their results to previously published work by using the same thresholds
and definitions. The `threshold` argument defines the spatial distance threshold
to establish a fusion event. The `n_min_length` argument defines the minimum
number of successive fixes that are required to establish a fusion event. The
`n_max_missing` defines the maximum number of allowable missed observations for
either individual in a dyad within a fusion event. The `allow_split` argument
defines if fusion events allow a temporary spatial splitting for one observation
without resulting in a fission event. Identification of fusion events has been
used to study resource seasonality [@Baden_2016], information transfer
[@Barocas_2016], decision making [@Merkle_2015], and habitat selection
[@Fortin_2009].

<!-- See Table Sup B: varying definitions of fission fusion 
TODO: mention from edge_dist
-->
<!-- TODO: add table numbers and cite them directly eg. Tab SB-3 instead of 
Supplement B: .... Table -->

```{r}
#| label: calc-fusion
#| eval: false
#| echo: true
fusion_events <- fusion_id(
  interindividual_dist,
  threshold = spatial_threshold, 
  n_min_length = 3, 
  n_max_missing = 1, 
  allow_split = TRUE
)
```


```{r}
#| label: fig-fusion
#| echo: false
#| eval: true
#| fig-cap: "A) Example locations for individuals 'A' and 'C' showing fusion event lasting three observations. B) A fusion event identified using fusion_id. Using the distance based edge-lists measured using edge_dist() and the dyad id returned by dyad_id(), fusion_id() a returns a column named 'fusionID' representing the fusion events defined with user specified constraints: the distance threshold ('threshold'), the minimum number of observations defining a fusion event ('n_min_length'), the maximum number of missing observations from either individual in a fusion event ('n_max_missing') and if a single observation where individuals are beyond the distance threshold should result in a fission event ('allow_split'). In B), a fusion event is identified for individuals A and C given they are within the distance threshold at timegroups 2, 3, and 5, and 'allow_split' is true."
tar_read(fig_fusion_events)
```


## Directional correlation delay

The directional correlation delay [@Nagy_2010], or the temporal delay at which
two individual's movement directions are most similar, can be measured using
`edge_delay()`. The function compares a focal individual's movement directions
at time $t$ to all of the movement direction's of a neighbor within a temporal
window to determine the temporal lag at which the neighbor's movement directions
are most similar to the focal individual. The directional correlation delay
ranges from negative values indicating the focal individual's direction is most
similar to the neighbor's previous movement direction to positive values
indicating the focal individual's direction is most similar to the neighbor's
future movement direction. The directional correlation delay can be aggregated
to determine mean directional correlation delay using `leader_edge_delay()`. The
mean directional correlation delay can be used to measure leadership patterns
and generate hierarchical leadership networks depending on the study system
[@Nagy_2010; @Flack_2013; @Nagy_2013; @Quera_2023; @Akos_2014].

```{r}
#| label: calc-edge-delay
#| eval: false
#| echo: true
delay <- edge_delay(
  edges = fusion_events,
  DT = DT,
  window = 3,
  id = id
)
```

```{r}
#| label: calc-leader-delay
#| eval: false
#| echo: true
leader_delay <- leader_edge_delay(
  delay,
  threshold = 0.5
)
```

```{r}
#| label: fig-edge-delay
#| eval: true
#| echo: false

#| fig-cap: "A) Example directional correlation delay for focal individual (ID1) 'C' and neighbour (ID2) 'G' indicating individual 'C' had a movement direction most similar to individual 'G' at the temporal delay of 0. Using the fusion events identified by fusion_id() and the 'direction' column returned by direction_step(), edge_delay() returns a column named 'direction_delay' representing the temporal delay at which the focal individual's movement direction is most similar to their neighbour's within a temporal window. B) Example movement directions for neighbour 'G' in the temporal window of timegroups 1-3 (set using argument 'window'). C) Aggregated directional delay using leader_edge_delay(). Using the 'direction_delay' column returned by edge_delay(), the leader_edge_delay() function returns a column named 'mean_direction_delay_delay' representing the mean directional correlation delay between two individuals and a column named 'mean_direction_delay' that represents the the mean correlation delay for each focal individual and all observed neighbours. Optionally, a threshold difference in direction can be used to subset the temporal delays included in the calculation of mean directional delay with the argument 'threshold'."
tar_read(fig_edge_delay)
```


# Discussion

- social networks, history of spatsoc
- high number of studies in review with IID, NND



# Conclusion

<!-- Test model only conclusions with increase species, increase systems given flexible powerful freely available functions -->

<!-- Association is only one layer of the animal social system -->

# Box 1 {.unnumbered}

Producer-scrounger dynamics describe two strategies of social foragers:
“producers” are those who discover patches and “scroungers” are those who join
patches discovered by others. This concept can be applied to animal telemetry
data by evaluating the position of individuals within groups, the individual’s
arrival timing to forage patches and characteristics of forage patches.
Individuals balance predation risk and foraging success by adjusting their
spatial position within groups  (Teichroeb 2015, Hirsch 2019), and the foraging
strategy selected by an individual can be influenced by the size of forage
patches (Hansen 2016).

Leader-follower patterns in groups of moving animals can be studied both to
determine factors that causally influence leadership, and for the influence of
leader-follower strategies on foraging success and downstream fitness. Applied
to animal telemetry, leadership can be defined simply by identifying the
individual at the foremost position along the front-to-back axis of the group’s
direction (Quera 2023, Harel 2021). Alternatively, leadership can be identified by
calculating the  directional correlation delay (Nagy 2010), the temporal delay
of shared directional movements between individuals. Leadership in a group of
moving animals can inform animal decision making, group coordination, and
migratory patterns (Flack 2018, Weesner 2023, Quera 2023, Kano 2021).

Individual traits can influence behaviors related to intragroup dynamics.
Dominance is related to foraging success (Caraco 1989, Teichroeb 2015) and
choice of foraging strategy (King 2009, Aplin and Morand-Ferron 2017, Lee 2016).
Dominant, aggressive, grooming and other interindividual behaviors are typically
measured using focal observations.  However, some metrics have been developed
for remotely estimating dominance behaviors. The directional correlation delay
can be used to derive hierarchies of leadership which can be a proxy for
dominance relationships in certain species (Flack 2013, Akos 2014). Movement
patterns that estimate dominance relationships can be identified from animal
telemetry including movement initiations (Amornbunchornvej 2018),
approach-avoidance behaviours (Strandburg-Peshkin 2015) and displacement events
(Evans 2018).



# References {.unnumbered}

:::{#refs}

:::


{{< include supplement/supplement-a.qmd >}}

{{< include supplement/supplement-b.qmd >}}

{{< include supplement/supplement-c.qmd >}}