# Supplement C: Selected Metrics {.unnumbered}


```{r}
#| include: false
targets::tar_source('R')
tar_load(taxized)
tar_load(geocoded)
tar_load(review)
tar_load(counted)
count_code_avail <- counted$count_code_avail
count_prog_lang <- counted$count_prog_lang
count_software <- counted$count_software
count_metric <- counted$count_metric
```

<!--
TODO:
- highlighting unavailability of selected metrics despite high use in literature
- detailed justification of not selected metrics
-->

## Directional correlation delay {.unnumbered}

```{r}
dirr_del <- count_metric[metric_agg == 'directional correlation delay']
```
 
Our scoping review found `r dirr_del[, uniqueN(covidence_number)]` studies that used 
the directional correlation delay metric, across 
`r taxized[covidence_number %in% dirr_del$covidence_number, uniqueN(species)]` 
study species. Only 
`r count_code_avail[covidence_number %in% dirr_del$covidence_number & analysis_code_availability == 'Yes', .N]` 
study had code available using a combination of 
MATLAB (closed source), CUDA (closed source), and Python (open source). At the
time when the scoping review was conducted, to the best of our knowledge, there
were no R packages that provide this metric.


## Position within group, rank position within group, time spent leading group, distance and direction to leader {.unnumbered}

```{r}
pos_wi <- count_metric[metric_agg %in% c('position within group', 
                                 'rank position within group',
                                 'time spent leading group',
                                 'distance to leader',
                                 'direction to leader')]
```

Our scoping review found `r pos_wi[, uniqueN(covidence_number)]` studies that used 
the position within group, rank position within group, time spent leading 
group metrics, distance to leader, or direction to leader metrics across 
`r taxized[covidence_number %in% pos_wi$covidence_number, uniqueN(species)]` 
study species. 
Only 
`r count_code_avail[covidence_number %in% pos_wi$covidence_number & analysis_code_availability == 'Yes', .N]`
study had code available. At the time when the scoping review was conducted, to
the best of our knowledge, there were no R packages that provide these metrics.


## Distance, direction and rank distance to group centroid {.unnumbered}

```{r}
to_grp_cent <- count_metric[metric_agg %in% c('direction to group centroid', 
                                      'distance to group centroid',
                                      'rank distance to group centroid',
                                      'group centroid')]
```

Our scoping review found `r to_grp_cent[, uniqueN(covidence_number)]` studies that used 
the group centroid, distance to group centroid, direction to group centroid, 
or rank distance to group centroid metrics, across 
`r taxized[covidence_number %in% to_grp_cent$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% to_grp_cent$covidence_number & analysis_code_availability == 'Yes', .N]`
studies had code available. The {swaRm} package provides functionality to 
calculate the group centroid and the distance to the group centroid. At the
time when the scoping review was conducted, to the best of our knowledge, there
were no R packages that provide the rank to distance to group centroid metric. 


## Polarization, directional alignment and interindividual direction {.unnumbered}


```{r}
align <- count_metric[metric_agg %in% c('polarization', 
                                'directional alignment',
                                'interindividual direction',
                                'nearest neighbour direction')]
```

Our scoping review found `r align[, uniqueN(covidence_number)]` studies that used 
the polarization, directional alignment or interindividual direction metrics, 
across
`r taxized[covidence_number %in% align$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% align$covidence_number & analysis_code_availability == 'Yes', .N]`
studies had code available. The  {CircStats} package provides functionality to 
calculate the polarization order parameter, as well as the {swaRm} package.
At the time when the scoping review was conducted, to the best of our knowledge,
there were no R packages that provide the directional alignment or
interindividual direction metrics.


## Behavioural zones {.unnumbered}


```{r}
zone <- count_metric[metric_agg %in% c('zones')]
```

Our scoping review found `r zone[, uniqueN(covidence_number)]` studies that used 
the behavioural zones metric, 
across
`r taxized[covidence_number %in% zone$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% zone$covidence_number & analysis_code_availability == 'Yes', .N]`
studies had code available. At the time when the scoping review was conducted,
to the best of our knowledge, there were no R packages that provide the
behavioural zones metric.


## Not selected {.unnumbered}

The following metrics were not selected due to their availability in 
existing R packages:

- {wildlifeDI}: DI, Don, Prox, Lixn, Cs, DI in direction, DI in displacement, 
HAI, IAB, Cr, Ca
- {corrMove}: diffusive movement correlation, drift movement correlation,
overall component movement correlation
- {changepoint}: change point analysis

The following metrics were not selected due to their availability in 
code availability accompanying publications:

- mFLICA provided in code accompanying @Amornbunchornvej_2018

The follow metrics were not selected due to the expected input data differing 
from existing {spatsoc} functions (XYT):

- metrics relating to posture