# Supplement D: Comparison of direction calculations {.unnumbered}

## Setup

### Load packages

```{r}
library(data.table)
library(units)
library(lwgeom)
library(sf)

library(spatsoc)

library(sftrack)
library(amt)
library(move2)
```

### Setup data

```{r}
#| results: "hide"
# {move2}
xy_move2 <- mt_read(mt_example())

# {amt}
data("amt_fisher")
xy_amt <- make_track(amt_fisher, x_, y_, t_)
coords_amt <- c('x_', 'y_')
xy_amt_sf <- st_as_sf(amt_fisher, coords = coords_amt, crs = 5070)

# {spatsoc}
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]
group_times(DT, 'datetime', '10 minutes')
crs <- 32736
splitBy <- NULL
timegroup <- 'timegroup'
threshold_sf <- units::as_units(50, 'm')
threshold <- 50
coords <- c('X', 'Y')
id <- 'ID'
source('../../spatsoc/R/get_sf.R')
xy_spatsoc <- get_sf(DT, coords = coords, crs = crs)

# {sftrack}
data(raccoon)
raccoon$timestamp <- as.POSIXct(as.POSIXlt(raccoon$timestamp, tz = "EST5EDT"))
xy_sftrack <- as_sftrack(
	data = raccoon,
	coords = c("longitude","latitude"),
	time = "timestamp",
	group = "animal_id",
	crs = 4326)
```

## Functionality

Comparing direction calculations in established packages, in terms of 
functionality provided. 

Spatial packages:

- {sf} does not include a direction function since it is not in the Simple Feature standard. 
- {lwgeom} calculates direction in the function `st_geod_azimuth` using the liblwgeom library. It does not accept NAs or empty geometries. It requires geographic coordinates. 

```{r}
#| error: TRUE
# Error: due to empty geometries
az <- st_geod_azimuth(xy_move2)
az <- st_geod_azimuth(xy_move2[!st_is_empty(xy_move2),])
head(az)
```

```{r}
#| error: TRUE
# Error: due to projected coordinates, not longlat
az <- st_geod_azimuth(xy_amt_sf)
az <- st_geod_azimuth(st_transform(xy_amt_sf, 4326))
head(az)
```
- {geosphere} calculates direction in [geosphere::bearing](https://github.com/rspatial/geosphere/blob/b215d4e569cc026903638b4653c27bdd7b2e1fb9/src/a_geodesic.cpp#L67-L77). It requires geographic coordinates. 

Applied packages:

- {move2} calculates direction in the function [`mt_azimuth`](https://gitlab.com/bartk/move2/-/blob/main/R/mt_azimuth.R?ref_type=heads) using `lwgeom::st_geod_azimuth` for geographic coordinates or `atan2` for projected coordinates
- {amt} calculates direction in the function `direction_abs` using `geosphere::bearing` for geographic coordinates and `atan2(diff_y, diff_x)` for projected coordinates
- {sftrack} calculates direction in the function [`step_direction`](https://github.com/mablab/sftrack/blob/HEAD/R/step_calculation.R#L143-L150) using `atan2` for projected coordinates and `geosphere::bearing` for geographic coordinates. 

All of {move2}, {amt} and {sftrack} calculate direction between a focal individual's locations (=step direction). None of these calculate the direction between an individual and something else - eg. a conspecific, or a group centroid.


## Results

Comparing direction calculations in established packages, in terms of 
results returned:

### {lwgeom} and `atan2`
```{r}
xy_spatsoc[, direction := c(lwgeom::st_geod_azimuth(geometry),
													 as_units(NA, 'rad'))]
xy_spatsoc[, direction_atan2 := c(atan2(diff(Y), diff(X)), NA)]

range(xy_spatsoc$direction, na.rm = TRUE)
range(xy_spatsoc$direction_atan2, na.rm = TRUE)
hist(drop_units(xy_spatsoc$direction) - xy_spatsoc$direction_atan2, 
     main = 'lwgeom::st_geod_azimuth with geographic - atan2 with projected')
```


### {sftrack}
```{r}
met_sftrack <- step_metrics(xy_sftrack)

xy_sftrack$geometry <- st_transform_proj(xy_sftrack$geometry, st_crs(26914))
met_sftrack_proj <- step_metrics(xy_sftrack)

range(met_sftrack$abs_angle, na.rm = TRUE)
range(met_sftrack_proj$abs_angle, na.rm = TRUE)
hist(met_sftrack$abs_angle - met_sftrack_proj$abs_angle, 
     main = 'sf_track with lonlat - projected')
```



## {move2}

```{r}
az <- mt_azimuth(dplyr::filter(xy_move2,!sf::st_is_empty(xy_move2)))
xy_move2_proj <- xy_move2
xy_move2_proj$geometry <- lwgeom::st_transform_proj(xy_move2$geometry, st_crs(32715))
st_crs(xy_move2_proj$geometry) <- NA
az_proj_but_na <- mt_azimuth(dplyr::filter(xy_move2_proj,!sf::st_is_empty(xy_move2)))

range(az, na.rm = TRUE)
range(az_proj_but_na, na.rm = TRUE)
hist(az - az_proj_but_na, main = 'move2 lonlat - projected but NA CRS')
# Note: move2 doesn't accept projected coordinates with non-NA CRS
```