# Supplement: Scoping review methods {#sec-scoping-methods}

We conducted a scoping review to identify metrics used in ecology studies for
measuring dyadic interactions, dominance hierarchies and leadership patterns. We
focused on metrics applicable to spatiotemporal data such as GPS telemetry. We
used two databases of academic literature: Web of Science and Scopus. We
established a test list of nine benchmark papers (Supplement A: Table 1) that
were relevant to our objectives and verified that each paper was indexed in our
selected databases. We used an iterative search method to fine-tune search terms
measuring the total number of papers returned and sensitivity of the search.
Sensitivity is defined as the proportion of potentially relevant papers returned
from the search estimated using the number benchmark papers returned from the
test list [@CollaborationforEnvironmentalEvidence_2022; 4.2.7]. Search terms
were established in four categories: topic, population, method and exclusion
(Supplement A: Table 2). Topic included terms associated with dyadic
interactions, dominance hierarchies and leadership patterns. Population included
terms associated with animals, wildlife and ecology. Method included terms
associated with data collection and data types. Finally, a list of exclusion
terms were used to remove clearly irrelevant papers. Each category of terms were
combined with OR operators and all three categories of terms were combined with
AND operators. The set of papers used in further steps was the combination of
the last iteration of search terms applied to both of the Web of Science (using
the “TS” field) and Scopus (using the “TITLE-ABS-KEY” field) databases. Only
articles between 1980-2023 were included using the PY (Web of Science) or
PUBYEAR (Scopus) fields. The resulting set of references were imported into
Covidence [@VeritasHealthInnovation_2025] for deduplication, title and abstract
screening, full text screening, and data extraction.

Before screening, articles were marked as duplicates using Covidence and manual
screening. Covidence detected 799 duplicates based on title, year, volume, and
authors, and manual screening identified an additional 128 duplicates. 

Articles were screened out based on the following criteria:

- Study does not describe dyadic or group metric related to leadership or
dominance (eg. @Papageorgiou_2020)
- Study describes metric that is not applicable to spatiotemporal data
(eg. @Allanic_2020 used focal observations and video recordings)
- Study does not describe metric beyond static interaction (eg. @Carter_2013 
used home range or spatial overlap)
- Study does not describe metric beyond group size (eg. @delMarDelgado_2021)
- Study full text is not available in English (eg. @Bernard_2002)
- Study is not a thesis 
- Study is not a book chapter (eg. @Bernstein_2019)

Despite a focus on animal ecology, some studies were discussed movement data
more generically or, alternatively, specifically focused on human or robotic
movement data (eg. @Long_2022). These studies were retained when the metrics
described were deemed applicable to animal spatiotemporal movement data.
Studies using mathematical models or simulations were largely beyond the
scope of this review but studies were retained when the parameters used as
inputs could be applied to spatial temporal data (eg. @Long_2015). Review
studies were read for building the scope of the introduction and discussion
but not included in data extraction unless they explicitly discussed 
dyadic measures (eg. @King_2018).

The goals of data extraction were to record the breadth of dyadic social and
dominance metrics applicable to spatiotemporal data, and the degree of code
availability. General information about each study was recorded including the
type of study (empirical, simulation, algorithm, model, or review), study
species and study region. Dyadic social metrics for each study were listed
including citations and interpretation for each metric used and data input
required. Dominance metrics derived from spatiotemporal data were listed
including citations and interpretation. Code availability was described
including location of code (eg. Zenodo, or GitHub), programming languages and
software packages used.

Metrics identified were aggregated using a manual process to identify synonymous
terms. This aggregation process was required due to variability in defining
metrics during data extraction, as well as differences between authors in
describing metrics (eg. distance to group "center", "centroid", or "center of
mass"). Each study was geocoded using the {tidygeocoder} package [@Cambon_2021]
and species were parsed using the {rgnparser} package [@Chamberlain_2023] which
provides the GNparser [@Mozzherin_2026] functionality in R. We calculated
frequency of data extracted for each categorical field including programming
language, software package, analysis code availability, geographic region, study
species and metric used. For many fields, authors did not provide sufficient to
determine how metrics were calculated, eg. not including the programming
language or software package used. In all cases, counts are presented as an
overall frequency including papers with missing fields.


```{r}
#| echo: false
#| label: tbl-form
#| tbl-cap: "Data extraction form for scoping review to identify metrics for studying intragroup dynamics"
data.table::fread('data/form.csv') |>
  tinytable::tt() |>
  style_tt(align = 'l')
```

```{r}
#| echo: false
#| label: tbl-benchmark
#| tbl-cap: "Benchmark papers used to evaluate iterative search results for scoping review"
targets::tar_read(benchmark_papers) |> 
  tinytable::tt() |>
  style_tt(align = 'l')
```

```{r}
#| echo: false
#| label: tbl-search-strings
#| tbl-cap: "Search strings for scoping review using Web of Science and Scopus"
targets::tar_read(search_strings) |> 
  tinytable::tt() |>
  style_tt(align = 'l')
```
