# Supplement: Scoping review results {#sec-scoping-results}

```{r}
#| include: false
targets::tar_source('R')

tar_load(review)
tar_load(geocoded)
tar_load(taxized)
tar_load(software_meta)

tar_load(counted)
count_code_avail <- counted$count_code_avail
count_prog_lang <- counted$count_prog_lang
count_software <- counted$count_software
count_metric <- counted$count_metric
count_dom_metric <- counted$count_dom_metric
```

## Results {.unnumbered}

PRISMA

-   3870 references imported for screening as 3870 studies
    -   128 duplicates identified manually
    -   799 duplicates identified by Covidence
-   2943 studies screened against title and abstract
    -   2527 studies excluded
-   416 studies assessed for full-text eligibility
    -   275 studies excluded
        -   184 Does not discuss dyadic measure
        -   86 Metric not applicable to spatiotemporal data
        -   2 Book chapter
        -   2 Full text not in English
        -   1 Thesis
    -   0 studies ongoing
    -   0 studies awaiting classification
-   141 studies included

::: {#suptbl-code-availability}
```{r}
#| echo: false
count_code_avail[, .N, analysis_code_availability][
  order(-N),
  .(
    `Code availability` = analysis_code_availability,
    N = paste_n_percent(N)
  )
] |>
  tt()
```

Count of code availability in references screened for full text extraction of scoping review
:::

```{r}
list_prog <- count_prog_lang[,
  uniqueN(covidence_number),
  .(prog = !is.na(programming_language))
]
unique_prog_lang <- count_prog_lang[,
  uniqueN(programming_language),
  covidence_number
]
```

Out of `r list_prog[, sum(V1)]` articles,
`r list_prog[!(prog), sum(V1)]` did not list a programming language or
GUI program. Of the `r list_prog[(prog), sum(V1)]` that did,
`r unique_prog_lang[V1 > 1, uniqueN(covidence_number)]` articles listed
more than one programming language or GUI program, therefore the total
count below will be greater than the number of articles.

::: {#suptbl-prog}
```{r}
#| echo: false
count_prog_lang[, .N, programming_language][
  order(-N),
  .(
    `Programming language` = programming_language,
    N = paste_n_percent(N)
  )
] |>
  tt()
```

Count of programming language or GUI program used in references screened for full text extraction of scoping review
:::

Out of `r count_software[, uniqueN(covidence_number)]` articles,
`r count_software[is.na(software_package_s_used), .N]` did not list a
software package. Of the
`r count_software[!is.na(software_package_s_used), uniqueN(covidence_number)]`
that did, `r count_software[, .N, covidence_number][N > 1, .N]` articles
listed more than one software package , therefore the total count below
will be greater than the number of articles.

:::{#suptbl-software}
```{r}
#| echo: false
#| label: tbl-software
count_software[, .N, software_package_s_used][
  order(-N),
  .(
    `Software package used` = software_package_s_used,
    N = paste_n_percent(N)
  )
] |>
  tt()
```

Count of software packages used in references screened for full text extraction of scoping review
:::

:::{#suptbl-temp}
```{r}
#| echo: false
ggplot(review) +
  geom_histogram(aes(yr), binwidth = 1) +
  labs(x = 'Year of publication', y = 'Count') +
  coord_cartesian(xlim = c(1980, 2024)) +
  theme_bw()
```

Count of references screened for full text extraction across temporal period (1980-2023) of scoping review
:::

:::{#suptbl-map}

```{r}
#| echo: false
countries <- ne_countries()

ggplot() +
  geom_sf(data = countries) +
  geom_hex(
    aes(long, lat),
    alpha = 1,
    bins = 20,
    data = geocoded[!is.na(long)]
  ) +
  coord_sf(crs = 54030, default_crs = 4326) +
  theme_void() +
  theme(legend.position = 'bottom')
```
Map of study regions georeferenced using {tidygeocoder} from references screened for full text extraction of scoping review
:::


:::{#suptbl-species}

```{r}
#| echo: false
taxized[, .N, parsed_species][
  order(-N),
  .(
    Species = parsed_species,
    N = paste_n_percent(N)
  )
] |>
  tt() |>
  style_tt(j = "Species", italic = TRUE)
```
Count of species parsed using {rgnparser} in references screened for full text extraction of scoping review
:::


:::{#suptbl-count-metrics}
```{r}
#| echo: false
count_metric[, .N, metric_agg][
  order(-N),
  .(
    `Metric` = metric_agg,
    N = paste_n_percent(N)
  )
] |>
  tt()
```
Count of metrics used to measure intragroup dynamics in references screened for full text extraction of scoping review
:::


:::{#suptbl-count-metrics-dominance}
```{r}
#| echo: false
count_dom_metric[, .N, dominance_metric_agg][
  order(-N),
  .(
    `Dominance metric` = dominance_metric_agg,
    N = paste_n_percent(N)
  )
] |>
  tt()
```
Count of metrics used to measure dominance in references screened for full text extraction of scoping review
:::


:::{#suptbl-defn-spattemp-groups}

```{r}
#| echo: false
review[definition_category == 'group', .(study_id, definition)] |>
  tt()
```
Review of definitions of spatiotemporal groups in references screened for full text extraction of scoping review
:::


:::{#suptbl-defn-leadership}
```{r}
#| echo: false
review[definition_category == 'leadership', .(study_id, definition)] |>
  tt()
```
Review of definitions of leadership in references screened for full text extraction of scoping review
:::


:::{#suptbl-defn-fission}
```{r}
#| echo: false
review[definition_category == 'fission fusion', .(study_id, definition)] |>
  tt()
```
Review of definitions of fission-fusion in references screened for full text extraction of scoping review
:::
