# Supplement B: Scoping Review Results {.unnumbered}

```{r}
#| include: false
targets::tar_source('R')

tar_load(review)
tar_load(geocoded)
tar_load(taxized)
tar_load(software_meta)

tar_load(counted)
count_code_avail <- counted$count_code_avail
count_prog_lang <- counted$count_prog_lang
count_software <- counted$count_software
count_metric <- counted$count_metric
count_dom_metric <- counted$count_dom_metric
```

## Results {.unnumbered}

### PRISMA {.unnumbered}

```
3870 references imported for screening as 3870 studies
  128 duplicates identified manually
  799 duplicates identified by Covidence
2943 studies screened against title and abstract
  2527 studies excluded
416 studies assessed for full-text eligibility
  262 studies excluded
    171  Does not discuss dyadic metric
    86  Metric not applicable to spatiotemporal data
    2  Book chapter
    2  Full text not in English
    1  Thesis
  0 studies ongoing
  0 studies awaiting classification
154 studies included
```


### Code availability {.unnumbered}

```{r}
count_code_avail[, .N, analysis_code_availability][order(-N), .(
  `Code availability` = analysis_code_availability,
  N = paste_n_percent(N)
)] |> tt()
```


### Programming language / GUI program {.unnumbered}

```{r}
list_prog <- count_prog_lang[, uniqueN(covidence_number), .(prog = !is.na(programming_language))]
unique_prog_lang <- count_prog_lang[, uniqueN(programming_language), covidence_number]
```

Out of `r list_prog[, sum(V1)]` articles, 
`r list_prog[!(prog), sum(V1)]` did not list a programming 
language or GUI program.  Of the 
`r list_prog[(prog), sum(V1)]` 
that did, `r unique_prog_lang[V1 > 1, uniqueN(covidence_number)]` 
articles listed more than one programming 
language or GUI program, therefore the total 
count below will be greater than the number of articles. 

```{r}
count_prog_lang[, .N, programming_language][order(-N), .(
  `Programming language` = programming_language,
  N = paste_n_percent(N)
)] |> tt()
```


### Software packages {.unnumbered}

Out of `r count_software[, uniqueN(covidence_number)]` articles, 
`r count_software[is.na(software_package_s_used), .N]` did not list a software package.
Of the 
`r count_software[!is.na(software_package_s_used), uniqueN(covidence_number)]` 
that did, `r count_software[, .N, covidence_number][N > 1, .N]` 
articles listed more than one software package 
, therefore the total 
count below will be greater than the number of articles. 

```{r}
count_software[, .N, software_package_s_used][order(-N), .(
  `Software package used` = software_package_s_used,
  N = paste_n_percent(N)
)] |> tt()
```


#### Software packages metadata {.unnumbered}

```{r}
software_meta |> 
  tt()
```


### Temporal {.unnumbered}

```{r}
ggplot(review) + 
  geom_histogram(aes(yr), binwidth = 1) + 
  labs(x = 'Year of publication', y = 'Count') + 
  coord_cartesian(xlim = c(1980, 2024)) +
  theme_bw()
```



### Geographic {.unnumbered}

```{r}
countries <- ne_countries()

crs <- 8857

ggplot() + 
  geom_sf(data = countries) + 
  geom_hex(aes(long, lat), alpha = 0.7, data = geocoded[!is.na(long)]) + 
  coord_sf(crs = crs, default_crs = 4326) + 
  theme_void() + 
  theme(legend.position = 'bottom')
```



### Species {.unnumbered}

```{r}
taxized[, .N, parsed_species][order(-N), .(
  Species = parsed_species,
  N = paste_n_percent(N)
)] |> tt() |> style_tt(j = "Species", italic = TRUE)
```


### Metrics {.unnumbered}

#### Definitions of spatiotemporal groups {.unnumbered}

```{r}
review[definition_category == 'group', .(study_id, definition)] |>
  tt()
```

#### Count of metrics measuring intragroup dynamics {.unnumbered}

```{r}
count_met <- count_metric[, .(N = uniqueN(covidence_number)), V1]

count_met[order(-N, V1), 
          .(`Metric` = V1, 
            N = paste_n_percent(N, count_metric[, uniqueN(covidence_number)]))] |> 
  tt()
```

### Definitions of leadership {.unnumbered}

```{r}
review[definition_category == 'leadership', .(study_id, definition)] |>
  tt()
```

### Definitions of fission fusion {.unnumbered}

```{r}
review[definition_category == 'fission fusion', .(study_id, definition)] |>
  tt()
```

#### Count of methods measuring dominance {.unnumbered}

```{r}
count_dom <- count_dom_metric[, .(N = uniqueN(covidence_number)), V1]

count_dom[order(-N, V1), 
          .(`Metric` = V1, 
            N = paste_n_percent(N, count_dom_metric[, uniqueN(covidence_number)]))] |> 
  tt()
```