---
title: "Methods"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



# Selecting metrics

Criteria: 

- highly cited, unavailable
- direct extension of available functions in spatsoc
- biology


Not selected:

- wildlifeDI
  - DI
  - Don
  - Prox
  - Lixn
  - Cs
  - DI in direction
  - DI in displacement
  - HAI
  - IAB
  - Cr
  - Ca
- corrMove
  - diffusive movement correlation
  - drift movement correlation
  - overall component movement correlation
- voronoi
- HMM
- change point detection


(mFLICA table 1)

- mFLICA 
  - available
- FLOCK
- LPD
- Granger causality
- Influence maximization


# Metrics selected, defined

For each

- Citation
- Interpretation 
- Formula
- Code example


## Directional correlation delay

```{r}
tar_load(count_metric)
tar_load(taxized)
tar_load(geocoded)
tar_load(review)

dirr_del <- count_metric[V1 == 'directional correlation delay']

dirr_del[, uniqueN(covidence_number)]

taxized[covidence_number %in% dirr_del$covidence_number, 
        unique(species)]

geocoded[covidence_number %in% dirr_del$covidence_number, 
        unique(region)]


review[grepl('directional correlation delay', metric_used_or_described), .(metric_used_or_described,
                                                                           citation_for_metric_if_not_introduced_in_this_current_paper, 
                                                                           interpretation)]
```





# Case study