---
title: "Methods"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



# Selecting metrics

Criteria: 

- highly cited, unavailable
- direct extension of available functions in spatsoc
- biology


Not selected:

- wildlifeDI
  - DI
  - Don
  - Prox
  - Lixn
  - Cs
  - DI in direction
  - DI in displacement
  - HAI
  - IAB
  - Cr
  - Ca
- corrMove
  - diffusive movement correlation
  - drift movement correlation
  - overall component movement correlation
- voronoi
- HMM
- change point detection


(mFLICA table 1)

- mFLICA 
  - available
- FLOCK
- LPD
- Granger causality
- Influence maximization


# Metrics selected, defined

For each

- Citation
- Interpretation, **emphasis on biology**
- Formula
- Code example
- Similar terms? eg. see metric synonyms


## Directional correlation delay

```{r}
tar_load(count_metric)
tar_load(taxized)
tar_load(geocoded)
tar_load(review)
count_code_avail <- tar_read(count_analysis_code_availability)

dirr_del <- count_metric[V1 == 'directional correlation delay']

# geocoded[covidence_number %in% dirr_del$covidence_number, 
#         unique(region)]

# review[grepl('directional correlation delay', metric_used_or_described), .(metric_used_or_described,
#                                                                            citation_for_metric_if_not_introduced_in_this_current_paper,
#                                                                            interpretation)]  |> tt()
```
 
Our review found `r dirr_del[, uniqueN(covidence_number)]` studies that used 
the directional correlation delay metric, across 
`r taxized[covidence_number %in% dirr_del$covidence_number, uniqueN(species)]` 
study species. Only 
`r count_code_avail[covidence_number %in% dirr_del$covidence_number & V1 == 'Yes', .N]` 
study had code available for 
MATLAB (closed source), CUDA (closed source), and Python (open source). To
our knowledge, there are no R packages that provide this metric. 



The directional correlation delay [@Nagy_2010] of individuals i, j is 

$$C_{ij} = [\overrightarrow{v_{i}}(t) * \overrightarrow{j}(t + \tau)]_{t}$$

where 

- $\overrightarrow{v_{i}}(t)$ is the normalized velocity of bird i at time t
- $\overrightarrow{v_{j}}(t + \tau)$ is the normalized velocity of bird j at time t + $\tau$
- Note that $C_{ij}(\tau)$ = $C_{ji}(-\tau)$
- Calculated only where pairs of individuals were less than 100 m apart


The maximum value of the directional correlation function $C_{ij}$ is at
$C_{ij}(\tau^{*}_{ij})$ where $\tau^{*}_{ij}$ is the directional correlation
delay time. $\tau^{*}_{ij}$ values focus on the relationship in pairs of
individuals, ignoring hierarchy changes caused by other individuals.

- Note that $\tau^{*}_{ij}$ = $-\tau^{*}_{ji}$
- Negative values indicate that flight directional changes of bird i fall behind that of bird j and therefore j is leading

Hierarchical networks can be generated using $\tau^{*}_{ij}$



## Position, rank position within group and time spent leading group

```{r}
pos_wi <- count_metric[V1 %in% c('position within group', 
                                 'rank position within group',
                                 'time spent leading group')]
```

Our review found `r pos_wi[, uniqueN(covidence_number)]` studies that used 
the position within group, rank position within group, or time spent
leading group metrics, across 
`r taxized[covidence_number %in% pos_wi$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% pos_wi$covidence_number & V1 == 'Yes', .N]`
studies had code available.  To our knowledge, there are no R packages that
provide this metric. 


- front-back distance along group's axis of movement [@Quera_2023]
- ordinal 1-6 position along group's traveling direction [@Burns_2012]
- relative positional ranks on the front-to-back axis were determined 
by multiplying their x–y locations by a rotation matrix based on the heading 
of the centroid [@Harel_2021]
- space transformation from geographical space to relative positions with respect
to group center and movement direction [@Andrienko_2013]


The position within group is the position along the front-to-back axis of the 
group's bearing. The rank position within group is the ordinal rank along
the front-to-back axis of the group's bearing. The time spent leading group
is an aggregate measure of the time ini the first rank position within group. 

- group = distance threshold
- at each step
- ?how to compare rank across varying sized groups


## Distance and direction to group centroid

```{r}
to_grp_cent <- count_metric[V1 %in% c('direction to group centroid', 
                                      'distance to group centroid',
                                      'rank distance to group centroid',
                                      'group centroid')]
```

Our review found `r to_grp_cent[, uniqueN(covidence_number)]` studies that used 
the group centroid, distance to group centroid, direction to group centroid, 
or rank distance to group centroid metrics, across 
`r taxized[covidence_number %in% to_grp_cent$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% to_grp_cent$covidence_number & V1 == 'Yes', .N]`
studies had code available. The {swaRm} package provides functionality to 
calculate the group centroid and the distance to the group centroid but not
to calculate the direction to group centroid or rank distance to group
centroid. 

The group centroid is the centroid of individual locations in a group. The
distance to group centroid is the distance between an individual's location
and the group centroid. The direction to group centroid is the absolute
bearing from an individual's location and the group centroid. The rank distance
to group centroid is the ordinal rank of individuals' distances from the 
group centroid. 


## Directional alignment

```{r}
align <- count_metric[V1 %in% c('polarization', 'directional alignment')]
```

Our review found `r align[, uniqueN(covidence_number)]` studies that used 
the directional alignment or polarization metrics, 
or rank distance to group centroid, across 
`r taxized[covidence_number %in% align$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% align$covidence_number & V1 == 'Yes', .N]`
studies had code available. The {swaRm} package provides functionality to 
calculate the polarization order parameter, as well as the {circStats} package.

Polarization measures the uniformity of bearings in a group of individuals 
[@Wang_2022] on a scale of 0-1 where values near 0 indicate that bearings point
in different directions and values near 1 indicate that bearings point in 
similar directions. 

```{r}
#| echo: true

# Random bearings (radians)
h <- runif(25, 0, 2 * pi)

library(swaRm)
pol_order(h)

library(CircStats)
r.test(h)

# Also interesting
circ.range(h)
```


# Case study

