---
title: "Case study"
date: "`r Sys.Date()`"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



# Case study

Functionality previously available in {spatsoc} includes spatiotemporal grouping,
edge list generation and data-stream permutations (Robitaille et al. 2019). 
These functions have allowed users to detect interactions between individuals, 
measure coocurrence within and across species, and generate social networks
from telemetry data. (TODO: cite examples citing spatsoc)

Spatiotemporal grouping can be performed given a telemetry dataset, 
temporal and spatial thresholds, and one of three grouping functions:
`group_pts`, `group_lines` or `group_polys`. Each return a unique identifier
for each spatiotemporal group, defined using point based distances, 
linear trajectory overlap or home range overlap, respectively. 


For example, the suggested `group_pts` workflow 
with example data from the package: 

```{r}
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = '20 minutes')

# Spatial grouping with timegroup
group_pts(DT, threshold = 5, id = 'ID',
          coords = c('X', 'Y'), timegroup = 'timegroup')

```

```{r}
#| echo: false
# Load packages
library(data.table)
suppressPackageStartupMessages(library(spatsoc))

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')
setorder(DT, timegroup)

# Spatial grouping with timegroup
DT <- group_pts(DT, threshold = 5, id = 'ID', 
                coords = c('X', 'Y'), timegroup = 'timegroup')

tinytable::tt(DT[order(timegroup)][1:5])
```

Users can also measure dyadic interindividual 
distance using the `edge_dist` function, and identify nearest neighbors using 
the `edge_nn` function. For example: 


```{r}
#| echo: true
#| eval: false

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')

# Interindividual distance with maximum distance threshold set at 50 m
interindividual_dist <- edge_dist(
  DT,
  threshold = 100,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)

# Identifying nearest neighbours, optionally returning interindividual distance
nearest_neighbours <- edge_nn(
  DT,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE
)
```

```{r}
#| echo: false
#| eval: true

library(data.table)
suppressPackageStartupMessages(library(spatsoc))

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')
setorder(DT, timegroup)

# Interindividual distance with maximum distance threshold set at 50 m
interindividual_dist <- edge_dist(
  DT,
  threshold = 100,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)
tinytable::tt(interindividual_dist[1:5])
```

```{r}
#| echo: false
#| eval: true

# Nearest neighbours, optionally returning interindividual distance
nearest_neighbours <- edge_nn(
  DT,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE
)
tinytable::tt(nearest_neighbours[1:5])
```

Building on these functions, {spatsoc}'s new functionality provides 
a finer understanding of an individual's behaviour with respect to conspecifics. 
We aim to provide users with accessible, flexible functions that will 
help them identify leader follower patterns, fission fusion dynamics and 
potentially dominance behaviours in their own animal telemetry data. 

## Position relative to groups

Extending {spatsoc}'s abilities to identify spatiotemporal groups, a new 
set of functions are now available to measure individuals position relative within
spatiotemporal groups. 

After identifying spatiotemporal groups with eg. `group_pts`, we 
can measure the group centroid.  

```{r}
#| eval: false
#| echo: true
xcol <- 'X'
ycol <- 'Y'
groupcol <- 'group'
group_centroid(DT, xcol, ycol, group = groupcol)
```

```{r}
#| eval: false
#| echo: true
xcol <- 'X'
ycol <- 'Y'
groupcol <- 'group'
group_pts(DT, threshold = 5, id = 'ID', 
                coords = c('X', 'Y'), timegroup = 'timegroup')
group_centroid(DT, xcol, ycol, group = groupcol)
```

Then we can measure each individual's direction and distance to the group centroid. 
TODO: insert refs with examples of biological interpretation of distance and direction to 
group centroid. 

```{r}
#| echo: true
#| eval: false
calc_dist_to_group_centroid(DT, xcol, ycol, return_rank = TRUE)
calc_dir_to_group_centroid(DT, xcol, ycol)
```

```{r}
#| echo: false
#| eval: true
g1 <- ggplot(DT) +
  geom_histogram(aes(dir_to_group_centroid), bins = 30) +
  labs(x = 'Direction to group centroid', y = '')
g2 <- ggplot(DT) +
  stat_halfeye(aes(dist_to_group_centroid, factor(rank_dist_to_group_centroid))) +
  labs(x = 'Distance to group centroid', y = 'Rank distance to group centroid')

g1 + g2
```

After considering how individuals are positioned relative to the group centroid, 
we can also measure how individuals are positioned relative to the average 
group direction. To do so, we rotate the coordinate system around the 
group centroid by the average bearing of all individuals in the group. Then 
we take the distance along this new axis to the measure of front-back position
within the group. TODO: refs where this is used, extended biological interpretation

```{r}
#| eval: false
#| echo: true
coords <- c('x', 'y')
projection <- 32636
calc_az(DT, coords, projection = projection)
position_within_group(DT, coords, return_rank = TRUE)
```

```{r}
#| eval: true
#| echo: false
coords <- c('X', 'Y')
projection <- 32636
setnames(DT, 'ID', 'id')
calc_az(DT, coords, projection = projection)
position_within_group(DT, coords, return_rank = TRUE)

sel_group <- DT[, .N, group][N > 3, sample(group, 1)]
sub_DT <- DT[group == sel_group]
slope <- sub_DT[1, tan(group_az)]
intercept <- sub_DT[1, group_mean_Y - slope * group_mean_X]
intercept_inv <- sub_DT[1, group_mean_Y - (-1/slope * group_mean_X)]

g_pos <- ggplot(sub_DT, aes(X, Y, color = id)) +
  geom_point(size = 0.8) +
  geom_text(aes(label = paste0(format(dist_along_group_az, digits = 1),
                               ' (', rank_dist_along_group_az, ')')),
                nudge_y = 2) +
  geom_point(color = 'black', aes(group_mean_X, group_mean_Y)) +
  geom_abline(slope = slope, intercept = intercept) +
  geom_abline(slope = -1/slope, intercept = intercept_inv,
              linewidth = 0.2) +
  theme_bw() +
  labs(x = '', y = '') +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  coord_fixed() +
  guides(color = 'none') +
  scale_x_continuous(expand = expansion(add = 10))

DT[, N_by_group := .N, group]
g_hist <- ggplot(DT[N_by_group > 1]) +
  geom_histogram(aes(dist_along_group_az), binwidth = 1) +
  labs(x = 'Distance along group az', y = '') +
  theme_bw()

g_hist2 <- ggplot(DT[N_by_group > 1]) +
  geom_histogram(aes(rank_dist_along_group_az), binwidth = 1) +
  labs(x = 'Rank distance along group az', y = '') +
  theme_bw()

g_dist <- ggplot(DT[N_by_group > 1]) +
  stat_halfeye(aes(dist_along_group_az, factor(rank_dist_along_group_az))) +
  labs(x = 'Distance along group az', y = 'Rank distance along group az') +
  theme_bw()

print(g_dist + ((g_hist / g_hist2) / g_pos))

```

# Case study

