---
title: "Case study"
date: "`r Sys.Date()`"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



# Case study

Functionality previously available in {spatsoc} includes spatiotemporal grouping,
edge list generation and data-stream permutations (Robitaille et al. 2019). 
These functions have allowed users to detect interactions between individuals, 
measure coocurrence within and across species, and generate social networks
from telemetry data. (TODO: cite examples citing spatsoc)

Spatiotemporal grouping can be performed given a telemetry dataset, 
temporal and spatial thresholds, and one of three grouping functions:
`group_pts`, `group_lines` or `group_polys`. Each return a unique identifier
for each spatiotemporal group, defined using point based distances, 
linear trajectory overlap or home range overlap, respectively. 


For example, the suggested `group_pts` workflow 
with example data from the package: 

```{r}
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = '20 minutes')

# Spatial grouping with timegroup
group_pts(DT, threshold = 5, id = 'ID',
          coords = c('X', 'Y'), timegroup = 'timegroup')

```

```{r}
#| echo: false
# Load packages
library(data.table)
suppressPackageStartupMessages(library(spatsoc))

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')
setorder(DT, timegroup)

# Spatial grouping with timegroup
DT <- group_pts(DT, threshold = 5, id = 'ID', 
                coords = c('X', 'Y'), timegroup = 'timegroup')

tinytable::tt(DT[order(timegroup)][1:5])
```

# Case study

