---
title: "Case study"
date: "`r Sys.Date()`"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



# Case study

Functionality previously available in {spatsoc} includes spatiotemporal grouping,
edge list generation and data-stream permutations (Robitaille et al. 2019). 
These functions have allowed users to detect interactions between individuals, 
measure coocurrence within and across species, and generate social networks
from telemetry data. (TODO: cite examples citing spatsoc)

Spatiotemporal grouping can be performed given a telemetry dataset, 
temporal and spatial thresholds, and one of three grouping functions:
`group_pts`, `group_lines` or `group_polys`. Each return a unique identifier
for each spatiotemporal group, defined using point based distances, 
linear trajectory overlap or home range overlap, respectively. 


For example, the suggested `group_pts` workflow 
with example data from the package: 

```{r}
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = '20 minutes')

# Spatial grouping with timegroup
group_pts(DT, threshold = 5, id = 'ID',
          coords = c('X', 'Y'), timegroup = 'timegroup')

```

```{r}
#| echo: false
# Load packages
library(data.table)
suppressPackageStartupMessages(library(spatsoc))

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')
setorder(DT, timegroup)

# Spatial grouping with timegroup
DT <- group_pts(DT, threshold = 5, id = 'ID', 
                coords = c('X', 'Y'), timegroup = 'timegroup')

tinytable::tt(DT[order(timegroup)][1:5])
```

Users can also measure dyadic interindividual 
distance using the `edge_dist` function, and identify nearest neighbors using 
the `edge_nn` function. For example: 


```{r}
#| echo: true
#| eval: false

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')

# Interindividual distance with maximum distance threshold set at 50 m
interindividual_dist <- edge_dist(
  DT,
  threshold = 100,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)

# Identifying nearest neighbours, optionally returning interindividual distance
nearest_neighbours <- edge_nn(
  DT,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE
)
```

```{r}
#| echo: false
#| eval: true

library(data.table)
suppressPackageStartupMessages(library(spatsoc))

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Temporal grouping
DT <- group_times(DT, datetime = 'datetime', threshold = '20 minutes')
setorder(DT, timegroup)

# Interindividual distance with maximum distance threshold set at 50 m
interindividual_dist <- edge_dist(
  DT,
  threshold = 100,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)
tinytable::tt(interindividual_dist[1:5])
```

```{r}
#| echo: false
#| eval: true

# Nearest neighbours, optionally returning interindividual distance
nearest_neighbours <- edge_nn(
  DT,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  returnDist = TRUE
)
tinytable::tt(nearest_neighbours[1:5])
```

Building on these functions, {spatsoc}'s new functionality provides 
a finer understanding of an individual's behaviour with respect to conspecifics. 
We aim to provide users with accessible, flexible functions that will 
help them identify leader follower patterns, fission fusion dynamics and 
potentially dominance behaviours in their own animal telemetry data. 

# Case study

