---
title: "Discussion"
date: "`r Sys.Date()`"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```



## Directional correlation delay

```{r}
tar_load(count_metric)
tar_load(taxized)
tar_load(geocoded)
tar_load(review)
count_code_avail <- tar_read(count_analysis_code_availability)

dirr_del <- count_metric[V1 == 'directional correlation delay']
```
 
Our review found `r dirr_del[, uniqueN(covidence_number)]` studies that used 
the directional correlation delay metric, across 
`r taxized[covidence_number %in% dirr_del$covidence_number, uniqueN(species)]` 
study species. Only 
`r count_code_avail[covidence_number %in% dirr_del$covidence_number & V1 == 'Yes', .N]` 
study had code available using a combination of 
MATLAB (closed source), CUDA (closed source), and Python (open source). To
our knowledge, there are no R packages that provide this metric. 


## Position within group, rank position within group, time spent leading group, distance and direction to leader

```{r}
pos_wi <- count_metric[V1 %in% c('position within group', 
                                 'rank position within group',
                                 'time spent leading group',
                                 'distance to leader',
                                 'direction to leader')]
```

Our review found `r pos_wi[, uniqueN(covidence_number)]` studies that used 
the position within group, rank position within group, time spent leading 
group metrics, distance to leader, or direction to leader metrics across 
`r taxized[covidence_number %in% pos_wi$covidence_number, uniqueN(species)]` 
study species. 
Only 
`r count_code_avail[covidence_number %in% pos_wi$covidence_number & V1 == 'Yes', .N]`
study had code available.  To our knowledge, there are no R packages that
provide these metrics. 


## Distance, direction and rank distance to group centroid

```{r}
to_grp_cent <- count_metric[V1 %in% c('direction to group centroid', 
                                      'distance to group centroid',
                                      'rank distance to group centroid',
                                      'group centroid')]
```

Our review found `r to_grp_cent[, uniqueN(covidence_number)]` studies that used 
the group centroid, distance to group centroid, direction to group centroid, 
or rank distance to group centroid metrics, across 
`r taxized[covidence_number %in% to_grp_cent$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% to_grp_cent$covidence_number & V1 == 'Yes', .N]`
studies had code available. The {swaRm} package provides functionality to 
calculate the group centroid and the distance to the group centroid. To our
knowledge, there are no R packages that provided the direction and
rank distance to group centroid metrics. 


## Polarization, directional alignment and interindividual direction


```{r}
align <- count_metric[V1 %in% c('polarization', 
                                'directional alignment',
                                'interindividual direction',
                                'nearest neighbour direction')]
```

Our review found `r align[, uniqueN(covidence_number)]` studies that used 
the polarization, directional alignment or interindividual direction metrics, 
across
`r taxized[covidence_number %in% align$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% align$covidence_number & V1 == 'Yes', .N]`
studies had code available. The {swaRm} package provides functionality to 
calculate the polarization order parameter, as well as the {CircStats} package.
To our knowledge, there are no R packages that provide the directional 
alignment or interindividual direction metrics. 


## Behavioural zones


```{r}
zone <- count_metric[V1 %in% c('zones')]
```

Our review found `r zone[, uniqueN(covidence_number)]` studies that used 
the behavioural zones metric, 
across
`r taxized[covidence_number %in% zone$covidence_number, uniqueN(species)]` 
study species. 
`r count_code_avail[covidence_number %in% zone$covidence_number & V1 == 'Yes', .N]`
studies had code available. To our knowledge, there are no R packages that
provided the behavioural zones metric.
- social networks, history of spatsoc
- high number of studies with IID, NND


Animal social dynamics have broad implications in ecology including resource
availability, disease transmission and collective behaviour. Social network
analysis can measure the structure of social systems, the consequences of an
individual's position in a network, the spread of disease or information across
networks, and the influence of the environment (Farine and Whitehead 2015). Open
source tools have been developed to derive social networks from animal telemetry
data (Robitaille et al 2019) using aggregated association rates or home range
overlap. The methods have been used to measure contact rates to measure social
cohesion (Bracken 2021), human-wildlife conflict (Boudreau 2021), socioecology
and resource availability (Peignier2019),  community structure (Sunga 2021),
influence of stress of social proximity (Keshavarzi 2023), social patterns in
non-social species (Heeres 2024).

