---
title: "Supplement A: Scoping Review Methods"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```


We conducted a scoping review to identify metrics used in ecology studies
focusing on dyadic interactions, dominance hierarchies and leadership patterns.
We focused on metrics applicable to spatiotemporal data such as GPS telemetry.
We used two databases of academic literature: Web of Science and Scopus. We
established a test list of nine benchmark papers (Supplement A: Table 1) that
were relevant to our objectives and verified that each paper was indexed in our
selected databases. We used an iterative search method to fine-tune search terms
measuring the total number of papers returned and sensitivity of the search.
Sensitivity is defined as the proportion of potentially relevant papers returned
from the search estimated using the number benchmark papers returned from the
test list (Env Evidence 4.2.7). Search terms were established in four
categories: topic, population, method and exclusion (Supplement A: Table 2).
Topic included terms associated with dyadic interactions, dominance hierarchies
and leadership patterns. Population included terms associated with animals,
wildlife and ecology. Method included terms associated with data collection and
data types. Finally, a list of exclusion terms were used to remove clearly
irrelevant papers. Each category of terms were combined with OR operators and
all three categories of terms were combined with AND operators. The set of
papers used in further steps was the combination of the last iteration of search
terms applied to both of the Web of Science (using the “TS” field) and Scopus
(using the “TITLE-ABS-KEY” field) databases. Only articles between 1980-2023
were included using the PY (Web of Science) or PUBYEAR (Scopus) fields. The
resulting set of references were imported into Covidence (Covidence systematic
review software) for deduplication, title and abstract screening, full text
screening, and data extraction.

Before screening, articles were marked as duplicates using Covidence and manual
screening. Covidence detected N duplicates based on title, year, volume, and
authors, and manual screening identified an additional N duplicates. 

Articles were screened out based on the following criteria:

- Study does not describe dyadic measure (eg. [@Papageorgiou_2020])
- Study describes metric that is not applicable to spatiotemporal data 
(eg. focal observations @Ballesta_2021)
- Study does not describe metric beyond static interaction (eg. home range or 
spatial overlap Carter_2013)
- Study does not describe metric beyond group size (eg. @Delgado_2021)
- Study full text is not available in English (eg. @Bernard_2002)
- Study is not a thesis (EXAMPLE)
- Study is not a book chapter (eg. @Bernstein_2019)

Despite a focus on animal ecology, some references were discussed movement data
more generically or, alternatively, specifically focused on human or robotic
movement data (EXAMPLE). These references were retained when the metrics
described were deemed applicable to animal spatiotemporal movement data.
References using mathematical models or simulations were largely beyond the
scope of this review but references were retained when the parameters used as
inputs could be applied to spatial temporal data (eg. @Long_2015). Review
references were retained for building the scope of introduction and discussion
but not included in data extraction (eg. @Herbert-Read_2016).

The goals of data extraction were to record the breadth of dyadic social and
dominance metrics applicable to spatiotemporal data, and the degree of code
availability. General information about each study was recorded including the
type of study (empirical, simulation, algorithm, model, or review), study
species and study region. Dyadic social metrics for each study were listed
including citations and interpretation for each metric used and data input
required. Dominance metrics derived from spatiotemporal data were listed
including citations and interpretation. Code availability was described
including location of code (eg. Zenodo, or GitHub), programming languages and
software packages used.

Metrics identified were aggregated using a manual process to identify synonymous
terms. This aggregation process was required due to variability in defining
metrics during data extraction, as well as differences between authors in
describing metrics (eg. distance to group "center", "centroid", or "center of mass"). 
Each study was geocoded using the {tidygeocoder}
package [@Cambon_2021] and species were parsed using the {rgnparser} package (CITE)
which provides the GNparser (CITE) functionality in R. We calculated frequency
of data extracted for each categorical field including programming language,
software package, analysis code availability, geographic region, study species
and metric used. For many fields, authors did not provide sufficient to
determine how metrics were calculated, eg. not including the programming
language or software package used. In all cases, counts are presented both as an
overall frequency including papers with missing fields and subset frequency
including only papers that provided sufficient information are retained.


## Form

Covidence #

Study ID

Title

Reviewer Name

Study information

- Type of study
  - Empirical
  - Simulation
  - Algorithm
  - Model
  - Review
  - Other
- Species
- Region

Dyadic social metric

- Metric used or described
- Software package(s) used
- Citation for metric if not introduced in this current paper
- Interpretation
- Data input required
- Definitions of "leader", "initiator", etc

Dominance

- Dominance metric
- Interpretation
- Citation for dominance metric if not introduced in this paper

Code availability

- Analysis code availability
- Programming language
- Analysis code link
- Analysis code availability location
- Other comments about code availability

Discussion points, etc


## Table A.1

```{r}
tar_source('R')
options("tinytable_print_output"="markdown")
tar_read(benchmark_papers) |> tt()
```

## Table A.2

```{r}
options("tinytable_print_output"=NULL)
tar_read(search_strings) |> tt()
```
