---
title: "Supplement B: Scoping Review Results"
date: "`r Sys.Date()`"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')

tar_load(review)
tar_load(geocoded)
tar_load(taxized)

tar_load(counted)
count_code_avail <- counted$count_analysis_code_availability
count_prog_lang <- counted$count_prog_lang
count_software <- counted$count_software
count_metric <- counted$count_metric
```

# Results

## PRISMA

```
3870 references imported for screening as 3870 studies
  128 duplicates identified manually
  799 duplicates identified by Covidence
2943 studies screened against title and abstract
  2527 studies excluded
416 studies assessed for full-text eligibility
  262 studies excluded
    171  Does not discuss dyadic metric
    86  Metric not applicable to spatiotemporal data
    2  Book chapter
    2  Full text not in English
    1  Thesis
  0 studies ongoing
  0 studies awaiting classification
154 studies included
```


## Code availability

```{r}
count_code_avail[, .N, V1][, .(
  `Code availability` = V1,
  N = paste_n_percent(N)
)] |> tt()
```




## Programming language / GUI program

Out of `r count_prog_lang[, uniqueN(covidence_number)]` articles, 
`r count_prog_lang[is.na(V1), .N]` did not list a programming 
language or GUI program.  Of the 
`r count_prog_lang[!is.na(V1), uniqueN(covidence_number)]` 
that did, `r count_prog_lang[, .N, covidence_number][N > 1, .N]` 
articles listed more than one programming 
language or GUI program, therefore the total 
count below will be greater than the number of articles. 


```{r}
count_prog <- count_prog_lang[!is.na(V1), .N, V1]

count_prog[order(-N), 
           .(`Programming language / GUI software` = V1, 
             N = paste_n_percent(N))] |> 
  tt()
```


## Software packages

Out of `r count_software[, uniqueN(covidence_number)]` articles, 
`r count_software[is.na(V1), .N]` did not list a software package.
Of the 
`r count_software[!is.na(V1), uniqueN(covidence_number)]` 
that did, `r count_software[, .N, covidence_number][N > 1, .N]` 
articles listed more than one software package 
, therefore the total 
count below will be greater than the number of articles. 

```{r}
count_soft_na_rm <- count_software[!is.na(V1), .N, V1]

count_soft_na_rm[order(-N),
           .(`Software package` = V1,
             N = paste_n_percent(N))] |>
  tt()
```


### Software packages metadata

```{r}
software_meta |> 
  tt()
```

## Metrics

```{r}
count_met <- count_metric[, .(N = uniqueN(covidence_number)), V1]

count_met[order(-N, V1), 
          .(`Metric` = V1, 
            N = paste_n_percent(N, count_metric[, uniqueN(covidence_number)]))] |> 
  tt()
```


### Not selected

- wildlifeDI (DI, Don, Prox, Lixn, Cs, DI in direction, DI in displacement, 
HAI, IAB, Cr, Ca, 
- corrMove (diffusive movement correlation, drift movement correlation,
overall component movement correlation)
- metrics relating to posture
- static interaction eg. home range overlap
- voronoi
- HMM
- change point detection
- mFLICA 
  - available
- FLOCK
- LPD
- Granger causality
- Influence maximization


## Temporal

```{r}
ggplot(review) + 
  geom_histogram(aes(yr), binwidth = 1) + 
  labs(x = 'Year of publication', y = 'Count') + 
  coord_cartesian(xlim = c(1980, 2024)) +
  theme_bw()
```



## Geographic

```{r}
countries <- ne_countries()

crs <- 8857

ggplot() + 
  geom_sf(data = countries) + 
  geom_hex(aes(long, lat), alpha = 0.7, data = geocoded[!is.na(long)]) + 
  coord_sf(crs = crs, default_crs = 4326) + 
  theme_void() + 
  theme(legend.position = 'bottom')
```



## Species

```{r}
#| fig-height: 9
ggplot(taxized) + 
  geom_bar(aes(y = fct_infreq(species)),
           fill = 'grey70',
           linewidth = 0.1) +
  theme_bw() + 
  labs(x = 'Count', y = 'Species')
```


## Fission fusion

```{r}
review[grepl('fission', metric_agg), .(study_id, definitions_of_leader_initiator_etc)] |> 
  tt()
```

