---
title: "Supplementary Materials B: Scoping Review Results"
toc: true
execute:
  echo: false
---

```{r}
#| include: false
targets::tar_source('R')
```

# Results

## PRISMA

(As of 2024-01-31)

```
3870 references imported for screening as 3870 studies
128 duplicates identified manually
799 duplicates identified by Covidence
2943 studies screened against title and abstract
2527 studies excluded
416 studies assessed for full-text eligibility
262 studies excluded
171  Does not discuss dyadic measure
86  Metric not applicable to spatiotemporal data
2  Book chapter
2  Full text not in English
1  Thesis
0 studies ongoing
0 studies awaiting classification
154 studies included
```


## Code availability

```{r}
tar_load(count_analysis_code_availability)

count_analysis_code_availability[, .N, V1][, .(
  `Code availability` = V1,
  N = paste_n_percent(N)
)] |> tt()
```




## Programming language / GUI program

```{r}
tar_load(count_programming_language)
```

Out of `r count_programming_language[, uniqueN(covidence_number)]` articles, 
`r count_programming_language[is.na(V1), .N]` did not list a programming 
language or GUI program.  Of the 
`r count_programming_language[!is.na(V1), uniqueN(covidence_number)]` 
that did, `r count_programming_language[, .N, covidence_number][N > 1, .N]` 
articles listed more than one programming 
language or GUI program, therefore the total 
count below will be greater than the number of articles. 


```{r}
count_prog <- count_programming_language[!is.na(V1), .N, V1]

count_prog[order(-N), 
           .(`Programming language / GUI software` = V1, 
             N = paste_n_percent(N))] |> 
  tt()
```


## Software packages

```{r}
tar_load(count_software)
tar_load(software_meta)
```


Out of `r count_software[, uniqueN(covidence_number)]` articles, 
`r count_software[is.na(V1), .N]` did not list a software package.
Of the 
`r count_software[!is.na(V1), uniqueN(covidence_number)]` 
that did, `r count_software[, .N, covidence_number][N > 1, .N]` 
articles listed more than one software package 
, therefore the total 
count below will be greater than the number of articles. 

```{r}
count_soft <- count_software[!is.na(V1), .N, V1]

count_soft[order(-N),
           .(`Software package` = V1,
             N = paste_n_percent(N))] |>
  tt()
```


### Software packages metadata

```{r}
software_meta |> 
  tt()
```

## Metrics

```{r}
tar_load(count_metric)
```

```{r}
count_met <- count_metric[, .(N = uniqueN(covidence_number)), V1]

count_met[order(-N, V1), 
          .(`Metric` = V1, 
            N = paste_n_percent(N, count_metric[, uniqueN(covidence_number)]))] |> 
  tt()
```


## Temporal

```{r}
#| include: false
tar_load(review)
```

```{r}
ggplot(review) + 
  geom_histogram(aes(yr), binwidth = 1) + 
  labs(x = 'Year of publication', y = 'Count') + 
  coord_cartesian(xlim = c(1980, 2024)) +
  theme_bw()
```



## Geographic

```{r}
tar_load(geocoded)
countries <- ne_countries()
```

```{r}
crs <- 8857

ggplot() + 
  geom_sf(data = countries) + 
  geom_hex(aes(long, lat), alpha = 0.7, data = geocoded[!is.na(long)]) + 
  coord_sf(crs = crs, default_crs = 4326) + 
  theme_void() + 
  theme(legend.position = 'bottom')
```



## Species

```{r}
tar_load(taxized)
```

```{r}
#| fig-height: 9
ggplot(taxized) + 
  geom_bar(aes(y = fct_infreq(species)),
           fill = 'grey70',
           linewidth = 0.1) +
  theme_bw() + 
  labs(x = 'Count', y = 'Species')
```
