---
title: "Home"
toc: true
---

```{r}
#| include: false
targets::tar_source('R')
```


## PRISMA

(As of 2024-01-31)

```
3870 references imported for screening as 3870 studies
	128 duplicates identified manually
	799 duplicates identified by Covidence
2943 studies screened against title and abstract
	2527 studies excluded
416 studies assessed for full-text eligibility
	262 studies excluded
		171  Does not discuss dyadic measure
		86  Metric not applicable to spatiotemporal data
		2  Book chapter
		2  Full text not in English
		1  Thesis
	0 studies ongoing
	0 studies awaiting classification
154 studies included
```


## Code availability

```{r}
#| echo: false
tar_load(count_analysis_code_availability)

count_analysis_code_availability[, .(
  `Code availability` = V1,
  N = paste_n_percent(N)
)]
```




## Programming language / software

```{r}
#| echo: false
tar_load(count_programming_language)

```

Out of `r count_programming_language[, uniqueN(covidence_number)]` articles, 
`r count_programming_language[is.na(V1), .N]` did not list a programming 
language or software. Of the 
`r count_programming_language[!is.na(V1), uniqueN(covidence_number)]` 
that did, `r count_programming_language[, .N, covidence_number][N > 1, .N]` 
articles listed more than one programming 
language or software, therefore the total 
count below will be greater than the number of articles. 


```{r}
#| echo: false
count_prog <- count_programming_language[!is.na(V1), .N, V1]

count_prog[order(-N), 
           .(`Programming language / software` = V1, 
             N = paste_n_percent(N))]
```