---
title: Extending {spatsoc} to measure intragroup social dynamics
author:
  - name: Alec L. Robitaille
    orcid: 0000-0002-4706-1762
    corresponding: true
    roles:
      - Conceptualization
      - Software
      - Investigation
      - Formal analysis
      - Data Curation
      - Writing - Original Draft
      - Writing - Review & Editing
      - Visualization
    affiliations:
      - Department of Biology, Memorial University of Newfoundland and Labrador
  - name: Quinn M. R. Webber
    orcid: 0000-0002-0434-9360
    corresponding: false
    # roles: 
    affiliations:
      - University of Guelph
  - name: Eric Vander Wal
    orcid: 0000-0002-8534-4317
    corresponding: false
    # roles: 
    affiliations:
      - Department of Biology, Memorial University of Newfoundland and Labrador
keywords:
  - R
  - animal social dynamics
  - spatiotemporal groups
  - social network analysis
  - collective movement
  - leader-follower
  - fission fusion
  - dominance hierarchies
abstract: | 
  1. Beyond proximity-based social networks and home range overlap, animal telemetry data can also be used to measure intragroup social dynamics including individual position within groups, individual and group level movement directions, leadership patterns and lagged follower behaviours.
  2. We used a scoping review of literature across domains, including behavioural ecology, collective movement, and GISciences, to identify widely used metrics for measuring intragroup social dynamics that are not openly available in the R programming language. 
  3. We present a case study illustrating 18 new functions for the R package {spatsoc} measuring intragroup social dynamics with animal telemetry data.
  4. The open availability of these new and flexible functions in {spatsoc} will allow researchers to easily measure intragroup social dynamics to more comprehensively measure the multifaceted animal social behaviours in their study systems. 
plain-language-summary: |
  New functions in the R package {spatsoc} to measure animal behaviour in social groups.
date: last-modified
bibliography: references/references.bib
csl: references/mee.csl
number-sections: true
---

# Introduction

```{r}
#| label: setup
#| include: false
targets::tar_source('_targets.R')
```

The R package {spatsoc} was developed to derive social networks from
animal telemetry data using aggregated association rates,
interindividual distances, and home range overlap [@Robitaille_2019].
These methods have been used to measure social cohesion [@Bracken_2022],
human-wildlife conflict [@Boudreau_2022], socioecology and resource
availability [@Peignier_2019], community structure [@Sunga_2021], the
influence of stress on social proximity [@Keshavarzi_2023], and social
patterns in non-social species [@Heeres_2024]. Despite these substantive
and varied applications of {spatsoc}, users lacked sufficient and
accessible functionality for measuring the behaviour of individuals
within social groups.

Many facets of group living need to be considered to better measure
animal social systems [@King_2018]. Beyond associations, there are
additional types of dyadic relationships including genetic, affiliative,
agonistic, and cooperative [@Farine_2015]. There are also interactions
between spatial phenotypes, social phenotypes, and environments
[@Webber_2023]. Using direct observations, researchers have studied
intragroup social dynamics including social foraging [@Barnard_1981;
@Giraldeau_1986; @Hirsch_2020], spatial position within groups
[@Krause_1994; @Monus_2008] and the influence of individual traits such
as age, sex, and dominance [@Furuichi_1983; @King_2009;
@Teichroeb_2015]. Unfortunately, direct observations can be
labour-intensive to collect and susceptible to observer bias
[@Smith_2021], with low spatial and temporal resolution making them
imperfect for studies of collective movement [@King_2018].

There are expanded opportunities in studying animal social systems by
pairing advances in remote tracking technology with emerging methods to
measure intragroup social dynamics. The availability and resolution of
remote tracking data continues to improve [@Kays_2015; @Kays_2022;
@Nathan_2022]. There are over two decades of research on measuring
interactions between moving individuals from the study of "moving point
objects" in GISciences [@Andersson_2008; @Dodge_2008; @Laube_2005;
@Miller_2021] to group coordination [@Couzin_2002; @Herbert-Read_2011],
leadership, and decision making [@King_2011; @Nagy_2010;
@Strandburg-Peshkin_2015]. Despite this, we suggest the lack of
availability of open source tools [@Hampton_2015; @Powers_2019;
@Wilson_2017] is the missing piece between the call to integrate
technology enhanced research with multidisciplinary approaches to better
study animal social systems [@Couzin_2023; @King_2018].

We conducted a scoping review to identify metrics that were applicable
to spatiotemporal data, and focused on dyadic interactions, leadership
patterns, and dominance hierarchies. Of the resulting metrics, we
selected those with a high evidence of use, and lack of availability in
the R programming language, either through R packages or supplemental
materials in the literature. Here we present
`r length(grep('cent|dir|dista|ali|del|zon|lead|get_geo|fus', ls('package:spatsoc')))`
new functions in the R package {spatsoc} [@Robitaille_2019] for
measuring intragroup dynamics ([Table @tbl-new-functions]). We will
demonstrate these new functions through a case study and discuss
biological interpretations.

# Methods

We conducted a scoping review to identify metrics for measuring
intragroup social dynamics including dyadic interactions, leadership
patterns, and dominance hierarchies. We focused on metrics applicable to
spatiotemporal data such as animal movement data. This scoping review
expands on the review conducted by @Joo_2018 by using interdisciplinary
sources (eg. @Miller_2021), including metrics measuring group-level
dynamics, leadership patterns, and lagged-follower behaviours, and by
focusing on the open source availability of these metrics in the R
programming language. See details scoping review methods in Supplement
[@sec-scoping-methods]. and Metrics were selected for inclusion in
{spatsoc} based on their relative importance as measured by frequency of
use in the review, and their lack of availability in R packages when the
review was conducted (Supplement [@sec-scoping-results]). Some metrics
were specifically not chosen for inclusion such as metrics relating to
posture, metrics relating to static interaction or home range overlap
and metrics based on statistical models such as Hidden Markov model and
Granger causality. A full list of metrics not selected is in Supplement
[@sec-selected].

```{r}
#| label: tbl-new-functions
#| echo: false
#| eval: true
#| tbl-cap: "Description of new functions in the R package {spatsoc} for measuring intragroup dynamics categorized by family. All avilable functions in the R package {spatsoc} including new and existing functions are listed in [Table @tbl-all-functions]."
functions <- fread('data/spatsoc-functions.tsv')
new_functions <- functions[Category == 'New']
new_functions[, .(Function, Description)] |>
  tt() |>
  group_tt(i = new_functions$Family) |>
  style_tt(align = 'l')
```

# Case study

Functions previously available in {spatsoc} can be used to identify
spatiotemporal grouping, generate distance based edge-lists and conduct
data-stream permutations (see existing functionality in
[@Robitaille_2019; @tbl-all-functions]. These functions have allowed
users to detect interactions between individuals, measure co-occurrence
within and across species, and generate social networks from animal
telemetry data [eg. @Albery_2025; @Bracken_2022; @Merkle_2024;
@Heeres_2024].

Spatiotemporal grouping can be performed given animal movement data,
temporal and spatial thresholds, and one of three grouping functions:
`group_pts()`, `group_lines()` or `group_polys()`. These return a unique
identifier for each spatiotemporal group defined by point-based
distances, linear trajectory overlap and home range overlap,
respectively.

`get_geometry()` is a new helper function added to {spatsoc} to provide
an alternative interface for providing coordinates and, optionally,
coordinate reference system. Instead of providing the names to columns
for the X and Y coordinates using the `coords` argument, the
`get_geometry()` function adds a simple feature geometry list column
(`sfc`) using the {sf} package [@Pebesma_2018]. The geometry method is
used by default for all {spatsoc} functions that require coordinates
when the `coords` argument is left null. This interface also extends
{spatsoc} to allow users to provide coordinates in any of the many
coordinate reference systems, both projected and longlat degrees, that
the {sf} package supports. The following examples and case study will
illustrate the geometry interface throughout.

For example, the suggested `group_pts()` workflow with the example data
from the package and the `get_geometry()` helper function to identify
spatiotemporal groups ([Table @tbl-group-pts]):

```{r}
#| label: calc-group-pts
#| eval: false
#| echo: true

# Load packages
library(data.table)
library(spatsoc)

# Read example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast the character column to POSIXct
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]

# Set variables
temporal_threshold <- '20 minutes'
spatial_threshold <- 50
id <- 'ID'
coords <- c('X', 'Y')
utm <- 32736

# Setup DT for geometry interface
get_geometry(DT, coords = coords, crs = utm)

# Temporal grouping
group_times(DT, datetime = 'datetime', threshold = temporal_threshold)

# Spatial grouping with timegroup
group_pts(
  DT,
  threshold = spatial_threshold,
  id = id,
  coords = coords,
  timegroup = 'timegroup'
)
```

```{r}
#| label: tbl-group-pts
#| tbl-cap: "Spatiotemporal groups measured using group_times() and group_pts(). group_times() returns a 'timegroup' column indicating temporal groups within the specified temporal threshold. group_pts() returns a 'group' column indicating spatiotemporal groups within the provided timegroup and specific distance threshold. Example output showing a subset of three spatiotemporal groups."
#| echo: false
DT <- tar_read(spatial_groups)
sel_groups <- DT[ID != 'A'][, .N, by = group][
  between(N, 2, 3),
  sample(group, 3)
]
tinytable::tt(DT[group %in% sel_groups][order(timegroup)][
  seq.int(6),
  .(ID, datetime, geometry, timegroup, group)
])
```

Users can also measure interindividual distance using the `edge_dist()`
function ([Table @tbl-edge-dist]) and identify nearest neighbours using
`edge_nn()` function ([Table @tbl-edge-nn]). For example:

```{r}
#| label: calc-edge-dist
#| echo: true
#| eval: false
# Measure interindividual distance
interindividual_dist <- edge_dist(
  DT,
  threshold = spatial_threshold,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE,
  fillNA = TRUE
)
```

```{r}
#| label: tbl-edge-dist
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_dist(). Using the 'timegroup' column returned by group_times(), edge_dist() returns a distance based edge-list where individuals in the column 'ID1' are within the provided distance threshold from 'ID2'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and NAs can be returned where an individual is not within the provided distance threshold if 'fillNA' is TRUE. Example output showing six observations of one timegroup with distances returned between focal individual 'ID1' and neighbour 'ID2'."
tar_load(distance_edges)
sel_tg <- distance_edges[, .N, by = .(timegroup, is.na(ID2))][, .N, timegroup][
  N > 1
][, sample(timegroup, 1)]
tinytable::tt(distance_edges[timegroup == sel_tg][seq.int(6)])
```

```{r}
#| label: calc-edge-nn
#| echo: true
#| eval: false
# Identify nearest neighbours
nearest_neighbours <- edge_nn(
  DT,
  id = id,
  timegroup = 'timegroup',
  returnDist = TRUE
)
```

```{r}
#| label: tbl-edge-nn
#| echo: false
#| eval: true
#| tbl-cap: "Distance based edge-lists measured using edge_nn(). Using the 'timegroup' column returned by group_times(), edge_nn() returns a distance based edge-list where individuals in the column 'NN' are the nearest neighbour to focal individuals in the column 'ID'. Optionally, measured distances can be returned in the column 'distance' if 'returnDist' is TRUE and a distance threshold can be provided using 'threshold'. Example output showing six observations of one timegroup with distances returned between focal individual 'ID' and nearest neighbour 'NN'."
tar_load(nn_edges)
sel_tg <- nn_edges[, .N, by = timegroup][N > 1][, sample(timegroup, 1)]
tinytable::tt(nn_edges[timegroup == sel_tg][seq.int(6)])
```

Building on these, {spatsoc}'s new functions provide a more detailed
measuring of an individual's behaviour with respect to their
conspecifics. We aim to provide users with accessible, flexible
functions that will help to identify leader-follower patterns, fission
fusion dynamics and potentially dominance behaviours from animal
telemetry data in their ecological systems.

## Position within groups

Extending {spatsoc}'s abilities to identify spatiotemporal groups, a new
set of functions are now available to measure individuals position
relative within spatiotemporal groups. After identifying spatiotemporal
groups with eg. `group_pts()`, we can measure the group centroid, as
defined by the mean of individual locations in a group, using
`centroid_group()` ([Table @tbl-cent-group]).

```{r}
#| label: calc-cent-group
#| eval: false
#| echo: true
# Measure group centroids
centroid_group(DT)
```

```{r}
#| label: tbl-cent-group
#| eval: true
#| echo: false
#| tbl-cap: "Group centroids measured using centroid_group(). Using the 'group' column returned by group_pts(), centroid_group() returns the column 'centroid' representing the centroid of all individual's locations in each spatiotemporal group. Example output showing a subset of the centroids for three spatiotemporal groups."
tar_load(group_centroids)
sel_groups <- group_centroids[ID != 'A'][, .N, by = group][
  between(N, 2, 3),
  sample(group, 3)
]
tinytable::tt(group_centroids[group %in% sel_groups][order(timegroup)][
  seq.int(6),
  .(ID, timegroup, group, geometry, centroid)
])
```

Then we can measure each individual's distance and direction to the
group centroid using `distance_to_centroid()` and
`direction_to_centroid()` ([Figure @fig-dist-dir-cent]). The distance to
group centroid is the geographic distance from the focal individual to
the group centroid. The direction to group centroid is the direction
from the focal individual to the group centroid. The rank of
individual's distances to the group centroid can be optionally returned
using the argument `return_rank`. Spatial position within groups with
respect to the group centroid has been studied in relation to predation
risk [@Heesen_2015; @Herbert-Read_2017], leadership [@Jolles_2017;
@Kano_2021] and group cohesion [@Harel_2021; @Bracken_2022].

```{r}
#| label: calc-dist-dir-cent
#| echo: true
#| eval: false
# Measure distance to group centroid
distance_to_centroid(DT, return_rank = TRUE)

# Measure direction to group centroid
direction_to_centroid(DT)
```

![A) Distance and B) direction to group centroids measured using distance_to_centroid() and direction_to_centroid(). Using the 'centroid' column returned by centroid_group, distance_to_centroid() and direction_to_centroid() return the columns 'distance_centroid' and 'direction_centroid' representing the focal individual's distance and direction to the group centroid. Optionally, the rank distance to the group centroid can be returned by distance_to_centroid() if 'return_rank' is true.](graphics/fig_dist_dir_cent.png){#fig-dist-dir-cent}


## Directions

Next, we can use movement directions to measure how individuals and
groups move together. We can measure an individual's movement direction
using `direction_step()` and the mean direction of each spatiotemporal
group identified by `group_pts()` using `direction_group()` ([Table
@tbl-dir-step-group]).

```{r}
#| label: calc-dir-step-group
#| echo: true
#| eval: false
# Measure movement direction
direction_step(DT, id = id)

# Measure mean movement direction of each spatiotemporal group
direction_group(DT)
```

```{r}
#| label: tbl-dir-step-group
#| echo: false
#| eval: true
#| warning: false
#| tbl-cap: "Individual and group mean directions measured using direction_step() and direction_group(). direction_step() returns the column 'direction' representing the movement direction of each individual to their next relocation. Using the spatiotemporal groups returned by group_pts, direction_group() returns the column 'group_direction' representing the mean direction of all individuals in each spatiotemporal group. Example output showing the movement directions of individuals and mean direction of two spatiotemporal groups."
tar_load(group_directions)
sel_groups <- group_directions[ID != 'A'][, .N, by = group][N == 3][, sample(
  group,
  2
)]
tinytable::tt(group_directions[ID != 'A' & group %in% sel_groups][order(
  timegroup
)][, .(ID, timegroup, group, direction, group_direction)])
```

We can measure polarization, the uniformity of movement directions in a
group of individuals, using `direction_polarization()` ([Figure
@fig-dir-pol]). Polarization is represented on a scale of 0-1 where
values near 0 indicate that individuals in a group are moving in
completely different directions (or "non-aligned") and values near 1
indicate that individuals are moving in similar directions or
("aligned"). Polarization in the movement direction of groups of animals
has been used to study cognition [@Wang_2022], learning
[@Vega-Trejo_2020], the influence of habitat [@Strandburg-Peshkin_2017]
and individual differences [@Jolles_2017] in collective behaviour.

```{r}
#| label: calc-dir-pol
#| echo: true
#| eval: false
# Measure direction polarization
direction_polarization(DT)
```

![Direction polarization measured using direction_polarization(). Using the 'direction' column returned by direction_step() and 'group' column returned by group_pts(), direction_polarization() returns a column named 'polarization' indicating the polarization in directions in each spatiotemporal group. Ranges of directions shown for a sample of groups in four equal interval categories of direction polarization. Values of polarization range from 0 (non-alignment) to 1 (alignment).](graphics/fig_dir_polarization.png){#fig-dir-pol}

The interindividual direction, or direction from a focal individual to a
neighbour, can be measured using `edge_direction()` ([Figure
@fig-edge-align-dir]). Like `edge_alignment()`, the interindividual
directions are returned by in an edge-list format. Interindividual
direction has been used to study risk exposure in schooling fish
[@Lemasson_2014], coordinated movement [@Herbert-Read_2016] and mass
migration [@Torney_2018].

```{r}
#| label: calc-edge-dir
#| echo: true
#| eval: false
# Measure interindividual direction
dyad_directions <- edge_direction(
  edges = interindividual_dist,
  DT = DT,
  id = id
)
```

The directional alignment, or relative difference between two
individuals' directions, can be measured using `edge_alignment()`
([Figure @fig-edge-align-dir]). The differences in movement directions
are returned in an edge-list format similar to `edge_dist()` and other
edge-list generating functions. Directional alignment has been used to
study leadership [@Sarova_2010], decision making
[@Strandburg-Peshkin_2015], and information transfer [@Sumpter_2018].

```{r}
#| label: calc-edge-align
#| echo: true
#| eval: false
# Measure directional alignment
directional_align <- edge_alignment(
  DT = DT,
  id = id,
  signed = FALSE
)
```

![A) Example locations for individuals 'E' (purple) and 'F' (blue) showing concurrent movement paths in timegroups 1, 2 and 3. B) Interindividual direction and directional_alignment() using edge_direction() and edge_alignment(). Using distance based edge-lists returned by edge_dist() or edge_nn() and the 'timegroup' column returned by group_times(), edge_direction() returns a column named 'direction_dyad' representing the direction between the focal individual ('ID1') and the neighbouring individual ('ID2'). Using the 'direction' column returned by direction_step(), the edge_alignment() function returns a column named 'direction_diff' representing the absolute different in movement directions between two individuals ('ID1' and 'ID2'). Optionally, the signed difference can be returned by edge_alignment() if the argument 'signed' is true.](graphics/fig_edge_dir_and_align.png){#fig-edge-align-dir}


## Behavioural zones

The behavioural zones [@Couzin_2002], or non-overlapping behavioural
regions around a focal individual, can measured using `edge_zones()`
([Figure @fig-edge-zones]). The distance from the focal individual to
each neighbour is used to assign the neighbours to behavioural zones. In
addition, the direction from the focal individual to each neighbour can
be used to determine if the neighbour is within a "blind volume" outside
of the individual's visual perception. @Couzin_2002 defined three
non-overlapping zones. The "zone of repulsion" is at a minimum distance
around the focal individual within which neighbours are expected to move
away to avoid collisions. The "zone of orientation" is around the focal
individual beyond the "zone of repulsion" within which neighbours are
expected to orient themselves to the movement of their neighbours. The
"zone of attraction" is beyond the "zone of orientation" where
neighbours are expected to be attracted to the position of the focal
individual. The `edge_zones()` function is designed to allow for
flexible definitions of zones including distance thresholds, zone labels
and blind volume thresholds. The behavioural zones metric has been used
to study the impact of individual movement speed and predator evasion
[@Klamser_2021] and identifying rules of interaction in large animal
groups [@Lukeman_2010].

```{r}
#| label: calc-edge-zones
#| eval: false
#| echo: true
# Define zone thresholds, labels and blind volume
zone_thresholds <- c(10, 25, 50)
zone_labels <- c('repulsion', 'orientation', 'attraction')
blind_volume <- 3

# Measure behavioural zones
zones <- edge_zones(
  dyad_directions,
  zone_thresholds = zone_thresholds,
  zone_labels = zone_labels,
  blind_volume = blind_volume
)
```

![Behavioural zones measured using edge_zones(). Using the 'distance' column returned by edge_dist(), the edge_zones() function returns the column 'zone' representing the behavioural zone of the neighbour (ID2) with respect to the focal individual (ID1). Optionally, the 'direction_dyad' column returned by edge_direction() can be used to define the blind volume, or the range of interindividual directions that are beyond the visual perception of the focal individual. Example shows positions of neighbour 'E' in behavioural zones with respect to focal individual 'G' across the sample period.](graphics/fig_edge_zones.png){#fig-edge-zones}

## Position relative to leader

After considering how individuals are positioned relative to the group
centroid, we can also measure how individuals are positioned relative to
the mean group direction. To do so, we first calculate each individual's
movement direction with `direction_step()` and the group's mean
direction with `direction_group()`. Then `leader_direction_group()`
rotates the coordinate system around the group centroid by the group's
mean direction. The distance along this new axis is the measure of
front-back position of an individual within the group [@Quera_2023;
@Harel_2021; @Andrienko_2013] ([Figure @fig-leader-dir-group]). The rank
of individual positions within spatiotemporal groups [@Burns_2012] can
be optionally returned when the argument 'return_rank' is true.

```{r}
#| label: calc-leader-dir-group
#| eval: false
#| echo: true
# Measure movement direction
direction_step(DT, id = id)

# Measure mean movement direction of each spatiotemporal group
direction_group(DT)

# Measure position relative to leader
leader_direction_group(DT, return_rank = TRUE)
```

![A) Example spatiotemporal group showing relative positions and movement directions of individuals (small points and grey arrows) with respect to the group centroid and mean group direction (large point and black arrow). B) Position along the group direction measured using leader_direction_group(). Using the 'direction' column returned by direction_step(), 'group_direction' column returned by direction_group() and 'centroid' column returned by centroid_group(), leader_direction_group() returns a column named 'position_group_direction' representing the front-back position along the mean group direction. Optionally, the rank position along the mean group direction if 'return_rank' is true.](graphics/fig_pos_group_dir.png){#fig-leader-dir-group}

Taking this simple, dynamic definition of leadership, we can identify
the distance and direction of individuals to the leader of each
spatiotemporal group using `distance_to_leader()` and
`direction_to_leader()` ([Figure @fig-dist-dir-lead]). The distance to
leader is the geographic distance from the focal individual to the
group's leader and the direction to leader is the direction from the
focal individual to the group's leader. The front-back position within
the group, and distance and direction to the group leader has been used
to study interaction rules [@Weesner_2023], leadership [@Quera_2023;
@Sarova_2010], risk avoidance [@Josephs_2016], and how reproduction
influences social structure [@Perez-Barberia_2018]. In addition, a user
can calculate an individual's time spent leading by calculating the rate
each individual was in the first ranked position out of the total
observations.

```{r}
#| label: calc-dist-dir-lead
#| eval: false
#| echo: true
# Measure distance to leader
distance_to_leader(DT)

# Measure direction to leader
direction_to_leader(DT)
```

![A) Distance and B) direction to group leaders measured using distance_to_leader and direction_to_leader. Using the 'rank_position_group_direction' column returned by leader_direction_group(), distance_to_leader() and direction_to_leader() return columns named 'distance_leader' and 'direction_leader' indicating the distance and direction of the focal individual to the leader of the spatiotemporal group. Note: for individuals identified as leader using 'position_group_direction', the distance is 0 and the direction is NA.](graphics/fig_dist_dir_leader.png){#fig-dist-dir-lead}

## Fusion events

Fusion events, or temporal periods where two individiuals are within a
given threshold distance from each other, can be identified using
`fusion_id()` ([Figure @fig-fusion]). Given the variability in defining
fission fusion dynamics in the literature ([@tbl-defn-fission]), the
`fusion_id()` function flexibly allows users to use specific thresholds
and definitions that match their study species and research questions.
This will also allow users to easily compare their results to previously
published work by using the same thresholds and definitions. The
`threshold` argument defines the spatial distance threshold to establish
a fusion event. The `n_min_length` argument defines the minimum number
of successive fixes that are required to establish a fusion event. The
`n_max_missing` defines the maximum number of allowable missed
observations for either individual in a dyad within a fusion event. The
`allow_split` argument defines if fusion events allow a temporary
spatial splitting for one observation without resulting in a fission
event. Identification of fusion events has been used to study resource
seasonality [@Baden_2016], information transfer [@Barocas_2016],
decision making [@Merkle_2015], and habitat selection [@Fortin_2009].

```{r}
#| label: calc-fusion
#| eval: false
#| echo: true
# Identify dyads
dyad_id(interindividual_dist, id1 = 'ID1', id2 = 'ID2')

# Identify fusion events
fusion_events <- fusion_id(
  interindividual_dist,
  threshold = spatial_threshold,
  n_min_length = 3,
  n_max_missing = 1,
  allow_split = TRUE
)

# Measure centroids of fusion events
centroid_fusion(
  interindividual_dist,
  DT,
  id = id
)
```

![A) Example fusion event for individuals 'J' and 'C' with black points representing the centroid of fusion observations returned by centroid_fusion(). B) A fusion event identified using fusion_id. Using the distance based edge-lists measured using edge_dist() and the dyad id returned by dyad_id(), fusion_id() a returns a column named 'fusionID' representing the fusion events defined with user specified constraints: the distance threshold ('threshold'), the minimum number of observations defining a fusion event ('n_min_length'), the maximum number of missing observations from either individual in a fusion event ('n_max_missing') and if a single observation where individuals are beyond the distance threshold should result in a fission event ('allow_split'). In B), a fusion event is identified for individuals J and C given they are within the distance threshold at timegroups 2, 3, and 4. Individual J is shown at timegroups 1 and 5 with no neighbours within the threshold spatial distance (ID2 is NA).](graphics/fig_fusion_events.png){#fig-fusion}

## Directional correlation delay

The directional correlation delay [@Nagy_2010], or the temporal delay at
which two individual's movement directions are most similar, can be
measured using `edge_delay()` ([Figure @fig-edge-delay]). The function
measures the similarity of movement directions of the focal individual
at time $t_{0}$ and the movement direction's of a neighbour within a
specified temporal window (eg. $t_{-2}, t_{-1}, t_{0}, t_{1}, t_{2}$).
The temporal delay is the difference in time where the neighbour's
movement directions are most similar to the focal individual's. The
directional correlation delay ranges from negative values indicating the
focal individual's direction is most similar to the neighbour's previous
movement direction to positive values indicating the focal individual's
direction is most similar to the neighbour's future movement direction.
The directional correlation delay can be aggregated to determine mean
directional correlation delay using `leader_edge_delay()`. The mean
directional correlation delay can be used to measure leadership patterns
and generate hierarchical leadership networks depending on the study
system [@Nagy_2010; @Flack_2013; @Nagy_2013; @Quera_2023; @Akos_2014].

```{r}
#| label: calc-edge-delay
#| eval: false
#| echo: true
# Measure directional correlation delay
delay <- edge_delay(
  edges = fusion_events,
  DT = DT,
  window = 3,
  id = id
)
```

```{r}
#| label: calc-leader-delay
#| eval: false
#| echo: true
# Measure mean directional correlation delay
leader_delay <- leader_edge_delay(
  delay,
  threshold = 0.5
)
```

![A) Example directional correlation delay for focal individual 'C' (ID1) and neighbour 'G' (ID2) indicating individual 'C' had a movement direction most similar to individual 'G' at the temporal delay of 0. Using the fusion events identified by fusion_id() and the 'direction' column returned by direction_step(), edge_delay() returns a column named 'direction_delay' representing the temporal delay at which the focal individual's movement direction is most similar to their neighbour's within a temporal window. B) Example movement directions for neighbour 'G' in the temporal window of timegroups 1-3 (set using argument 'window'). C) Aggregated directional delay using leader_edge_delay(). Using the 'direction_delay' column returned by edge_delay(), the leader_edge_delay() function returns a column named 'mean_direction_delay_delay' representing the mean directional correlation delay between two individuals and a column named 'mean_direction_delay' that represents the the mean correlation delay for each focal individual and all observed neighbours. Optionally, a threshold difference in direction can be used to subset the temporal delays included in the calculation of mean directional delay with the argument 'threshold'.](graphics/fig_edge_delay.png){#fig-edge-delay}

# Implications

The open source package {spatsoc} [@Robitaille_2019] for the R
programming language [@RCoreTeam_2025a] has improved accessibility to
methods for spatiotemporal grouping and social network analysis with
animal telemetry data [eg. @Albery_2025; @Bracken_2022; @Merkle_2024;
@Heeres_2024]. Despite this, proximity based associations are only one
dimension and studying intragroup social dynamics will allow researchers
to further measure animal social systems. The new functions in {spatsoc}
and the accompanying case study presented here make openly available a
series of metrics that while commonly used in the literature have been
largely absent from R packages. Paired with the continued development of
animal tracking technology [@Nathan_2022; @Beardsworth_2022] and novel
data collection methods [@Koger_2023; @Pedrazzi_2025], the open
availability of these metrics in R will broaden their application across
research questions, including long-standing ecological concepts such as:

-   Social foraging theory and producer-scrounger dynamics where
    “producers” are those who discover patches and “scroungers” are
    those who join patches discovered by others [@Barnard_1981;
    @Vickery_1991; @Giraldeau_2000]. This concept can be applied to
    animal telemetry data by evaluating the position of individuals
    within groups, the individual’s arrival timing to forage patches and
    the characteristics of forage patches. Individuals balance predation
    risk and foraging success by adjusting their spatial position within
    groups [@Teichroeb_2015; @Hirsch_2020], and the foraging strategy
    selected by an individual can be influenced by the size of forage
    patches [@Hansen_2016].

-   Leader-follower patterns in groups of moving animals can be studied
    both to determine factors that causally influence leadership, and
    for the influence of leader-follower strategies on foraging success
    and downstream fitness. Applied to animal telemetry data, leadership
    can be defined simply by identifying the individual at the foremost
    position along the front-to-back axis of the group’s direction
    [@Quera_2023; @Harel_2021] or by calculating the directional
    correlation delay between individuals and building leadership
    hierarchies [@Nagy_2010]. Leadership in a group of moving animals
    can inform animal decision making, group coordination, and migratory
    patterns [@Flack_2018; @Weesner_2023; @Quera_2023; @Kano_2021].

-   Individual traits can influence behaviours related to intragroup
    dynamics. Dominance is related to foraging success [@Caraco_1989;
    @Teichroeb_2015] and choice of foraging strategy [@King_2009;
    @Aplin_2017; @Lee_2016]. Dominant, aggressive, grooming and other
    interindividual behaviours are typically measured using focal
    observations. However, some metrics have been developed for remotely
    estimating dominance behaviours. The directional correlation delay
    can be used to derive hierarchies of leadership which can be a proxy
    for dominance relationships in certain species [@Flack_2013;
    @Akos_2014]. Dominance relationships can also be inferred from
    animal telemetry data using movement initiations
    [@Amornbunchornvej_2018], approach-avoidance behaviours
    [@Strandburg-Peshkin_2015] and displacement events [@Evans_2018].

The lack of open source code and transparency of methods alongside
research ([@tbl-code-availability], [@tbl-prog], [@tbl-software])
continues to slow the progress of science by limiting the broader
application of new methods. We hope that the availability of these
metrics in {spatsoc} will allow researchers to apply them to diverse
study species, systems or research questions, including novel
applications that could have management and conservation implications
for vulnerable species [@Ramos_2021].

# Open science

All code used to produce this manuscript including figures and tables
are available on GitHub at LINK <!-- TODO: github link --> and on
Zeonodo at LINK <!-- TODO: github link -->. The data used in this case
study are included with the package and can be imported with:

```{r}
#| eval: false
#| echo: true
library(spatsoc)

library(data.table)

DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))
```

The code for producing the scoping review results, figures, tables and
manuscript was developed as a reproducible pipeline with the R package
{targets} [@Landau_2021b]. Figures and tables were constructed using
{ggplot2} [@Wickham_2016a], {ggdist} [@Kay_2025], {patchwork}
[@Pedersen_2025] and {tintytable} [@Arel-Bundock_2025]. The manuscript
was produced using {quarto} [@Allaire_2024]. The {spatsoc} package
gratefully depends on the R packages {adehabitatHR} [@Calenge_2024],
{data.table} [@Barrett_2025], {igraph} [@Csardi_2026], {sf}
[@Pebesma_2018], {lwgeom} [@Pebesma_2025], {CircStats} [@Lund_2025],
{units} [@Pebesma_2016], and {rlang} [@Henry_2026].

# Acknowledgements

We thank Edzer Pebesma and Dewey Dunnington for support with bug reports
and feature requests for the {sf}, {s2} and {lwgeom} packages related to
the development of this new functionality in {spatsoc}. We thank Martin
Leclerc, Rick Heeres and Shane Frank for testing the package's new
functionality ahead of its release. We thank members of the Wildlife
Evolutionary Ecology Lab for their comments on previous versions of this
manuscript. We thank the rOpenSci organisation including staff, editors,
reviewers and contributors for their continued efforts to foster a
community around open data, software and reproducibility.

<!-- TODO: add funding, comps?-->

# References {.unnumbered}

::: {#refs}
:::

{{< include supplement/supplement-a.qmd >}}

{{< include supplement/supplement-b.qmd >}}

{{< include supplement/supplement-c.qmd >}}